---
title: "Generalized linear models"
---

## Model

$$\begin{align}
g(\mathbb{E}(Y|X)) &= \beta_0 + \beta_1 x_1 + \dots + \beta_k x_k \\
\Leftrightarrow g(\mu) &= \beta^\top X
\end{align}$$

There are three components to a GLM:

1.  **Systematic component**: The right-hand side is a linear combination of predictors: $\beta_0 + \beta_1 x_1 + \dots + \beta_k x_k$.

2.  **Random component**: The outcome $Y$ on the left-hand side, must follow a distribution from the **exponential family**.

3.  **Link function**: $Y$ is wrapped in a link function $g(Y)$ that relates the **mean** $\mathbb{E}(Y)$ to the linear predictors.

## Canonical link function

A **canonical link function** is the link function that directly connects the mean $\mu$ of a distribution to its natural (canonical) parameter $\theta$ in the exponential family form. Which means we want:

$$g(\mu) = \beta^\top X = \theta$$

For instance, suppose $Y$ follows a Poisson distribution. We know from the exponential‐family representation @tbl-exp-fam-distr, that the canonical parameter $\theta = \log(\lambda)$, where $\lambda = \mu$ because $\mathbb{E}(Y) = \lambda$ for Poisson.

Therefore, a **log link** function is the canonical link for a Poisson distribution:

$$g(\mu) = \log(\mu) = \log(\lambda) = \theta$$

Likewise, you can ask ChatGPT to prove that a **logit link** is the canonical link if $Y$ follows a Bernoulli, or Binomial distribution, and many other cases.

## Poisson regression

Poisson regression models are generalized linear models with the logarithm as the link function.

## Offset

rate = count / unit time

offset converts a count to rate

```{r}
week <- 1:7
amount_time <- c(0.85, 1.1, 0.85, 1.2, 0.95, 1.1, 0.9)
count <- c(90, 76, 37, 27, 19, 13, 9)
rate <- count / amount_time
df <- data.frame(week, amount_time, count, rate)
```

$$\log(\text{count}) = \beta_0 + \beta \cdot \text{week}$$

If we are fitting $\text{rate}$ instead of $\text{count}$:

$$\begin{align}
\log(\text{rate}) = \log\left(\frac{\text{count}}{\text{amount time}}\right) &= \beta_0 + \beta \cdot \text{week} \\
\Leftrightarrow \log(\text{count}) - \log(\text{amount time}) &= \beta_0 + \beta \cdot \text{week} \\
\Leftrightarrow \log(\text{count}) &= \beta_0 + \beta \cdot \text{week} + \log(\text{amount time})
\end{align}$$

The $\log(\text{amount time})$ is an offset.

```{r}
library(ggplot2)
```

```{r}
mod <- glm(count ~ week + offset(log(amount_time)), family = poisson)
mod
```

```{r}
pred <- predict(mod, type = "response")
ggplot(df, aes(x = week, y = count)) +
  geom_point() +
  geom_line(aes(y = pred)) +
  theme_classic()
```
