---
title: "Generalized linear models"
---

## Model

$$g(Y) = \beta_0 + \beta_1 x_1 + \dots + \beta_k x_k$$

There are three components to a GLM:

- Random component: the probability distribution of the outcome Y, any distribution within the **exponential family**.

- Systematic component: the linear combinations of predictors $\beta_0 + \beta_1 x_1 + \dots + \beta_k x_k$

- Link function $g(Y)$: the link between random and systematic components.

## Common link functions



## Offset

rate = count / unit time

offset converts a count to rate

## Poisson regression

Poisson regression models are generalized linear models with the logarithm as the link function.

```{r}
week <- 1:7
amount_time <- c(0.85, 1.1, 0.85, 1.2, 0.95, 1.1, 0.9)
count <- c(90, 76, 37, 27, 19, 13, 9)
rate <- count / amount_time
df <- data.frame(week, amount_time, count, rate)
```


$$\log(\text{count}) = \beta_0 + \beta \cdot \text{week}$$

If we are fitting $\text{rate}$ instead of $\text{count}$:

$$\begin{align}
\log(\text{rate}) = \log\left(\frac{\text{count}}{\text{amount time}}\right) &= \beta_0 + \beta \cdot \text{week} \\
\Leftrightarrow \log(\text{count}) - \log(\text{amount time}) &= \beta_0 + \beta \cdot \text{week} \\
\Leftrightarrow \log(\text{count}) &= \beta_0 + \beta \cdot \text{week} + \log(\text{amount time})
\end{align}$$

The $\log(\text{amount time})$ is an offset.

```{r}
library(ggplot2)
```

```{r}
mod <- glm(count ~ week + offset(log(amount_time)), family = poisson)
mod
```

```{r}
pred <- predict(mod, type = "response")
ggplot(df, aes(x = week, y = count)) +
  geom_point() +
  geom_line(aes(y = pred)) +
  theme_classic()
```

