# Least squares estimation

::: {.callout-note}

## Preparation

Download influenza incidence data here:

{{< downloadthis data/flu_inc.rds dname="flu_inc" label="flu_inc.rds" type=light >}}

:::

## Methods

Refer to @def-residual, the residual sum of squares (RSS) is the sum of the squares of residuals.

$$RSS = \sum_{i = 1}^{n}[y_i - f(x_i)]^2$$

-   $y_i$: the observed value.
-   $f(x_i)$: the predicted value.
-   $y_i - f(x_i)$: the residual.

The least squares method finds the optimal parameters by **minimizing the RSS**.

## Code {#sec-lse-code}

```{r, message=FALSE, warning=FALSE}
library(tidyr)
library(deSolve)
library(ggplot2)
library(RColorBrewer)
```

We will use data collected during a H1N1 influenza outbreak in an elementary school in Pennsylvania [@cauchemez2011]. Let load the data and have a look.

```{r}
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "100%"
df <- readRDS("data/flu_inc.rds")

ggplot(df, aes(x = day, y = inc)) +
  geom_point() +
  labs(x = "Day", y = "Incidence") +
  theme_minimal()
```

We will fit a deterministic SEIR model and reuse the code from @lst-seir-det. Let plug some initial values for the parameters and run the model.

```{r, echo=FALSE}
seir_mod <- function(beta, sigma, gamma, S0, E0, I0, R0, times) {
  
  equations <- function(time, variables, parameters) {
    with(as.list(c(variables, parameters)), {
      dS <-   -beta * I * S
      dE <-    beta * I * S - sigma * E
      dI <-    sigma * E - gamma * I
      dR <-    gamma * I
      dCInc <- sigma * E
      return(list(c(dS, dE, dI, dR, dCInc)))
    })
  }
  
  parameters_values <- c(beta = beta, sigma = sigma, gamma = gamma)
  initial_values <- c(S = S0, E = E0, I = I0, R = R0, CInc = 0)
  
  out <- ode(y = initial_values, times = times, func = equations, parms = parameters_values)

  out <- as.data.frame(out)
  # Compute incidence from cumulative incidence
  out$Inc <- c(I0, diff(out$CInc))
  out$CInc <- NULL
  out
}
```

Looking at the original paper, the school has 370 students [@cauchemez2011]. So we set the initial values as below.

```{r}
S0 <- 369
E0 <- 0
I0 <- 1
R0 <- 0
```

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "100%"
pred <- seir_mod(beta = 0.004, sigma = 0.5, gamma = 0.5, S0 = S0, E0 = E0, I0 = I0, R0 = R0, times = df$day)

df_plot <- pivot_longer(pred, cols = S:Inc, names_to = "comp", values_to = "n")

my_palette <- brewer.pal(11, "PuOr")[c(10, 1, 4, 3, 8)]

ggplot(df_plot, aes(x = time, y = n, color = comp)) +
  geom_point() + geom_line() +
  scale_color_manual(values = my_palette, breaks = c("S", "E", "Inc", "I", "R")) +
  labs(color = NULL, y = NULL, x = "Day") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Now just plot the incidence to compare with our data.

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "100%"
df_plot <- pred[,c("time", "Inc")]
ggplot(df_plot, aes(x = time, y = Inc)) +
  geom_point(color = my_palette[3]) +
  geom_line(color = my_palette[3]) +
  geom_point(data = df, aes(x = day, y = inc)) +
  labs(x = "Day", y = "Incidence") +
  theme_minimal()
```

We need to calculate the residual sum of squares. The `rss()` function must be defined this way with 2 arguments `data` and `par` so that it could be passed to the `optim()` function.

```{r}
rss <- function(data, par) {
  pred <- seir_mod(beta = par[1], sigma = par[2], gamma = par[3], S0 = S0, E0 = E0, I0 = I0, R0 = R0, times = 0:(length(data) - 1))
  sum((data - pred$Inc)^2)
}
```

The `optim()` function needs at least 2 inputs:

- `par`: a vector of initial values for the parameters to be optimized over.
- `fn`: a function to be minimized.
- `...`: other arguments will be passed to `fn`, here our function `rss()` needs an additional argument `data`, so we pass `data = df$inc`.

```{r}
seir_optim <- optim(par = c(0.004, 0.5, 0.5), fn = rss, data = df$inc)
seir_optim
```

Let plug these values into the model and compare with our data.

::: {#fig-lse-seir}

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "100%"
pred <- seir_mod(beta = seir_optim$par[1], sigma = seir_optim$par[2], gamma = seir_optim$par[3], S0 = S0, E0 = E0, I0 = I0, R0 = R0, times = df$day)
df_plot <- pred[,c("time", "Inc")]
ggplot(df_plot, aes(x = time, y = Inc)) +
  geom_point(color = my_palette[3]) +
  geom_line(color = my_palette[3]) +
  geom_point(data = df, aes(x = day, y = inc)) +
  labs(x = "Day", y = "Incidence") +
  theme_minimal()
```

Deterministic SEIR model fitted with least squares estimation.

:::

Since we are using a "closed population" SEIR model, $R_0 = \frac{\beta}{\gamma} S_0$.

::: {#lst-lse-r0}

```{r}
rnum0 <- seir_optim$par[1] * S0 / seir_optim$par[3]
rnum0
```

$R_0$ from deterministic SEIR model fitted with least squares estimation.

:::

This is quite close to 1.4 (95% CI: 1.2, 1.5) as reported by [@cauchemez2011], considering that they used more complicated methods.