---
title: "Model fitting"
execute:
  echo: false
  warning: false
---

```{r}
#| results: false
library(odin)
library(bbmle)

# The observed data points
df <- read.csv("data/flu_inc.csv")

# A model to fit
odin_seir <- odin::odin({
  # Derivatives
  deriv(S) <- -beta * S * I
  deriv(E) <- beta * S * I - sigma * E
  deriv(I) <- sigma * E - gamma * I
  deriv(R) <- gamma * I
  deriv(CInc) <- sigma * E

  # Initial conditions
  initial(S) <- S_init
  initial(E) <- E_init
  initial(I) <- I_init
  initial(R) <- R_init
  initial(CInc) <- CInc_init

  # Parameters and initial values
  beta <- user(0.004)
  sigma <- user(0.5)
  gamma <- user(0.5)
  S_init <- user(369)
  E_init <- user(0)
  I_init <- user(1)
  R_init <- user(0)
  CInc_init <- user(0)
})

seir_mod <- function(beta, sigma, gamma, S0, E0, I0, R0, times) {
  # Set values for a new run
  odin_run <- odin_seir$new(beta = beta, sigma = sigma, gamma = gamma, S_init = S0, E_init = E0, I_init = I0, R_init = R0)

  # Run the model
  out <- data.frame(odin_run$run(times))

  # Compute incidence from cumulative incidence
  out$Inc <- c(I0, diff(out$CInc))
  out$CInc <- NULL
  out
}

S0 <- 369
E0 <- 0
I0 <- 1
R0 <- 0

mll <- function(beta, sigma, gamma, sd_inc) {
  # Make sure that parameters are positive
  beta <- exp(beta)
  sigma <- exp(sigma)
  gamma <- exp(gamma)
  sd_inc <- exp(sd_inc)

  pred <- seir_mod(beta = beta, sigma = sigma, gamma = gamma, S0 = S0, E0 = E0, I0 = I0, R0 = R0, times = 0:(length(df$inc) - 1))
  pred <- pred$Inc
  # Return the negative log-likelihood
  - sum(dnorm(x = df$inc, mean = pred, sd = sd_inc, log = T))
}

starting_param_val <- list(beta = 0.004, sigma = 0.5, gamma = 0.5, sd_inc = 3)
estimates <- mle2(minuslogl = mll, start = lapply(starting_param_val, log), method = "Nelder-Mead")
params <- exp(coef(estimates))

optim_df <- data.frame(
  beta  = as.numeric(params[1]),
  sigma = as.numeric(params[2]),
  gamma = as.numeric(params[3]),
  sd    = as.numeric(params[4])
)

# Poisson likelihood: incidence counts given predicted mean λ = predicted incidence
mll_pois <- function(beta, sigma, gamma) {
  beta  <- exp(beta)
  sigma <- exp(sigma)
  gamma <- exp(gamma)
  pred <- seir_mod(beta = beta, sigma = sigma, gamma = gamma, S0 = S0, E0 = E0, I0 = I0, R0 = R0, times = 0:(length(df$inc) - 1))$Inc
  pred <- pmax(pred, 1e-9)  # guard against log(0)
  -sum(dpois(x = df$inc, lambda = pred, log = TRUE))
}

estimates_pois <- mle2(minuslogl = mll_pois,
                       start = list(beta = log(0.004), sigma = log(0.5), gamma = log(0.5)),
                       method = "Nelder-Mead")
params_pois <- exp(coef(estimates_pois))

optim_pois_df <- data.frame(
  beta  = as.numeric(params_pois[1]),
  sigma = as.numeric(params_pois[2]),
  gamma = as.numeric(params_pois[3])
)
```

Model fitting, or **calibration**, is the process of **adjusting** a model's **parameters** so that the **model's predictions** align **as closely as possible** with **real-world observations**.

In simple terms, you run the model with a given set of parameters and compare its simulated predictions with observed data points. Consider the example of an SEIR model with the following differential equations:

$$\frac{dS}{dt} = -\beta SI$$

$$\frac{dE}{dt} = \beta SI - \sigma E$$

$$\frac{dI}{dt} = \sigma E - \gamma I$$

$$\frac{dR}{dt} = \gamma I$$

Below you'll find an interactive plot. The dots represent observed data points, and the line shows the model's predicted incidence cases. Drag the sliders for $\beta$, $\sigma$ and $\gamma$ and watch how the curve responds, aiming to pull it through the dots. The readout tells you how well the current parameters score - smaller is better on both the residual sum of squares (RSS) and the negative log-likelihood ($-\log L$). Click **Optimize** to jump to the values a computer picks.

```{r}
ojs_define(ojsdata = df, optim_df = optim_df, optim_pois_df = optim_pois_df)
```

```{ojs}
import { createSlider, createButton, injectStyle, styleMathLabel } from "./_slider.js"
d3 = require("d3@7")
optim = transpose(optim_df)[0]
optim_pois = transpose(optim_pois_df)[0]
```

```{ojs}
//| echo: false
// Shared SEIR simulator (RK4) used by all three illustrations.
mf_sim = {
  const S0 = 369, E0 = 0, I0 = 1, R0 = 0;

  function step(state, beta, sigma, gamma, dt) {
    const [S, E, I, R, C] = state;
    const d = (S, E, I, R) => [
      -beta * S * I,
       beta * S * I - sigma * E,
       sigma * E - gamma * I,
       gamma * I,
       sigma * E
    ];
    const k1 = d(S, E, I, R);
    const k2 = d(S + 0.5*dt*k1[0], E + 0.5*dt*k1[1], I + 0.5*dt*k1[2], R + 0.5*dt*k1[3]);
    const k3 = d(S + 0.5*dt*k2[0], E + 0.5*dt*k2[1], I + 0.5*dt*k2[2], R + 0.5*dt*k2[3]);
    const k4 = d(S +     dt*k3[0], E +     dt*k3[1], I +     dt*k3[2], R +     dt*k3[3]);
    return [
      S + dt*(k1[0] + 2*k2[0] + 2*k3[0] + k4[0])/6,
      E + dt*(k1[1] + 2*k2[1] + 2*k3[1] + k4[1])/6,
      I + dt*(k1[2] + 2*k2[2] + 2*k3[2] + k4[2])/6,
      R + dt*(k1[3] + 2*k2[3] + 2*k3[3] + k4[3])/6,
      C + dt*(k1[4] + 2*k2[4] + 2*k3[4] + k4[4])/6
    ];
  }

  // Returns daily incidence for days 0..days (inclusive).
  // Day 0 is I0; day k (k>=1) is cumulative incidence from (k-1) to k.
  function run(beta, sigma, gamma, days) {
    const dt = 0.02;
    const perDay = Math.round(1 / dt);
    let state = [S0, E0, I0, R0, 0];
    const cInc = new Array(days + 1);
    cInc[0] = 0;
    for (let d = 1; d <= days; d++) {
      for (let k = 0; k < perDay; k++) state = step(state, beta, sigma, gamma, dt);
      cInc[d] = state[4];
    }
    const inc = new Array(days + 1);
    inc[0] = I0;
    for (let d = 1; d <= days; d++) inc[d] = cInc[d] - cInc[d - 1];
    return inc;
  }

  return { run, S0 };
}
```

```{ojs}
//| echo: false

viewof mf_explore = {
  const data = transpose(ojsdata).filter(d => d.day <= 30);
  const DAYS = 30;

  const wrapper = document.createElement("div");
  wrapper.style.cssText = "font-family:system-ui,-apple-system,sans-serif;max-width:760px;margin:0 auto;";
  wrapper.appendChild(injectStyle());

  // ── Sliders ──
  const SL = {};
  SL.beta  = createSlider("β",  0.001, 0.010, 0.0001, 0.004, "#3b82f6", "blue");
  SL.sigma = createSlider("σ",  0.3,   3.0,   0.05,   0.5,   "#7c3aed", "purple");
  SL.gamma = createSlider("γ",  0.3,   3.0,   0.05,   0.5,   "#0891b2", "teal");
  styleMathLabel(SL.beta, SL.sigma, SL.gamma);

  const slRow = document.createElement("div");
  slRow.style.cssText = "display:flex;gap:20px;margin-bottom:12px;";
  slRow.appendChild(SL.beta.el);
  slRow.appendChild(SL.sigma.el);
  slRow.appendChild(SL.gamma.el);
  wrapper.appendChild(slRow);

  // ── Readout row ──
  const readRow = document.createElement("div");
  readRow.style.cssText = "display:flex;gap:20px;margin-bottom:10px;align-items:baseline;font-family:'SF Mono',SFMono-Regular,Menlo,Consolas,monospace;";
  const mkStat = (label, color) => {
    const box = document.createElement("div");
    box.style.cssText = "display:flex;flex-direction:column;gap:2px;flex:1;";
    const l = document.createElement("span");
    l.style.cssText = "font-size:11px;font-weight:600;color:#64748b;text-transform:uppercase;letter-spacing:0.5px;font-family:system-ui,-apple-system,sans-serif;";
    l.textContent = label;
    const v = document.createElement("span");
    v.style.cssText = `font-size:22px;font-weight:800;color:${color};`;
    box.appendChild(l);
    box.appendChild(v);
    return { el: box, set: t => v.textContent = t };
  };
  const rssOut = mkStat("Residual sum of squares", "#d97706");
  const nllOut = mkStat("Negative log-likelihood", "#7c3aed");
  readRow.appendChild(rssOut.el);
  readRow.appendChild(nllOut.el);
  wrapper.appendChild(readRow);

  // ── Plot ──
  const W = 720, H = 300;
  const M = { top: 12, right: 16, bottom: 34, left: 42 };
  const innerW = W - M.left - M.right;
  const innerH = H - M.top - M.bottom;

  const svg = d3.create("svg")
    .attr("viewBox", `0 0 ${W} ${H}`)
    .style("width", "100%")
    .style("max-width", `${W}px`)
    .style("height", "auto")
    .style("display", "block")
    .style("background", "#fff");

  const xSc = d3.scaleLinear().domain([0, DAYS]).range([0, innerW]);
  const ySc = d3.scaleLinear().domain([0, 16]).range([innerH, 0]);

  const g = svg.append("g").attr("transform", `translate(${M.left},${M.top})`);

  // Grid
  g.append("g")
    .attr("stroke", "#e2e8f0").attr("stroke-dasharray", "2,3")
    .selectAll("line.yg").data(ySc.ticks(5)).join("line").attr("class","yg")
    .attr("x1", 0).attr("x2", innerW).attr("y1", d => ySc(d)).attr("y2", d => ySc(d));

  // Axes
  g.append("g").attr("transform", `translate(0,${innerH})`)
    .call(d3.axisBottom(xSc).ticks(7))
    .attr("font-size", 11);
  g.append("g")
    .call(d3.axisLeft(ySc).ticks(5))
    .attr("font-size", 11);
  g.append("text")
    .attr("x", innerW/2).attr("y", innerH + 28)
    .attr("text-anchor", "middle")
    .style("font-size", "12px").style("fill", "#475569")
    .text("Day");
  g.append("text")
    .attr("transform", `translate(-32,${innerH/2}) rotate(-90)`)
    .attr("text-anchor", "middle")
    .style("font-size", "12px").style("fill", "#475569")
    .text("Incidence");

  // Residual segments (gray, behind line)
  const gResid = g.append("g");

  // Fit line
  const linePath = g.append("path")
    .attr("fill", "none")
    .attr("stroke", "#d97706").attr("stroke-width", 2.5);

  // Observed dots (on top)
  g.selectAll("circle.obs")
    .data(data).join("circle").attr("class", "obs")
    .attr("cx", d => xSc(d.day))
    .attr("cy", d => ySc(d.inc))
    .attr("r", 3.2)
    .attr("fill", "#0f172a");

  const lineGen = d3.line().x(d => xSc(d.t)).y(d => ySc(d.inc));

  // ── Buttons ──
  const btnRow = document.createElement("div");
  btnRow.style.cssText = "display:flex;gap:10px;margin-top:8px;max-width:320px;";
  const optBtn = createButton("▶  Optimize", "auto");
  const resetBtn = createButton("↻  Reset", "reset");
  btnRow.appendChild(optBtn.el);
  btnRow.appendChild(resetBtn.el);
  wrapper.appendChild(svg.node());
  wrapper.appendChild(btnRow);

  // ── Update ──
  function update() {
    const beta = SL.beta.val(), sigma = SL.sigma.val(), gamma = SL.gamma.val();
    const inc = mf_sim.run(beta, sigma, gamma, DAYS);
    const pred = inc.map((v, t) => ({ t, inc: v }));
    linePath.attr("d", lineGen(pred));

    // Residuals and scores (over observed days only)
    const sdHat = optim.sd;
    let rss = 0, nll = 0;
    const resSegs = [];
    for (const d of data) {
      const p = inc[d.day];
      const r = d.inc - p;
      rss += r * r;
      // Normal log-likelihood with estimated sd
      nll += 0.5*Math.log(2*Math.PI*sdHat*sdHat) + (r*r)/(2*sdHat*sdHat);
      resSegs.push({ day: d.day, yObs: d.inc, yPred: p });
    }

    gResid.selectAll("line.r")
      .data(resSegs).join("line").attr("class", "r")
      .attr("x1", d => xSc(d.day)).attr("x2", d => xSc(d.day))
      .attr("y1", d => ySc(d.yObs)).attr("y2", d => ySc(d.yPred))
      .attr("stroke", "#94a3b8").attr("stroke-width", 1.2)
      .attr("stroke-dasharray", "2,2");

    rssOut.set(rss.toFixed(1));
    nllOut.set(nll.toFixed(1));
  }

  SL.beta.input.addEventListener("input",  () => { SL.beta.sync();  update(); });
  SL.sigma.input.addEventListener("input", () => { SL.sigma.sync(); update(); });
  SL.gamma.input.addEventListener("input", () => { SL.gamma.sync(); update(); });

  // Animate to target values
  function animateTo(target, ms = 600) {
    const start = { beta: SL.beta.val(), sigma: SL.sigma.val(), gamma: SL.gamma.val() };
    const t0 = performance.now();
    function frame(now) {
      const k = Math.min(1, (now - t0) / ms);
      const e = k < 0.5 ? 2*k*k : 1 - Math.pow(-2*k + 2, 2)/2; // easeInOutQuad
      SL.beta.update( start.beta  + (target.beta  - start.beta)  * e);
      SL.sigma.update(start.sigma + (target.sigma - start.sigma) * e);
      SL.gamma.update(start.gamma + (target.gamma - start.gamma) * e);
      update();
      if (k < 1) requestAnimationFrame(frame);
    }
    requestAnimationFrame(frame);
  }

  optBtn.el.addEventListener("click", () =>
    animateTo({ beta: optim.beta, sigma: optim.sigma, gamma: optim.gamma }));
  resetBtn.el.addEventListener("click", () =>
    animateTo({ beta: 0.004, sigma: 0.5, gamma: 0.5 }));

  update();
  return wrapper;
}
```

## Methods

Model fitting methods typically include [@rui2024]:

- Least squares estimation (LSE)
- Maximum likelihood estimation (MLE):
  - Optimisation: Nelder-Mead
  - Simulated annealing
  - MCMC
  - Newton Raphson
- Bayesian methods

### Least squares estimation {#sec-calib-lse}

For each day, the **residual** is the gap between the data point and the model's prediction. Least squares penalises *squared* residuals, so a single large miss is much worse than many small ones. The blue squares below have side length equal to the residual — their total area is the RSS. Slide the parameters until the squares shrink.

```{ojs}
//| echo: false

viewof mf_lse = {
  const data = transpose(ojsdata).filter(d => d.day <= 30);
  const DAYS = 30;

  const wrapper = document.createElement("div");
  wrapper.style.cssText = "font-family:system-ui,-apple-system,sans-serif;max-width:760px;margin:0 auto;";
  wrapper.appendChild(injectStyle());

  // ── Sliders ──
  const SL = {};
  SL.beta  = createSlider("β",  0.001, 0.010, 0.0001, 0.004, "#3b82f6", "blue");
  SL.sigma = createSlider("σ",  0.3,   3.0,   0.05,   0.5,   "#7c3aed", "purple");
  SL.gamma = createSlider("γ",  0.3,   3.0,   0.05,   0.5,   "#0891b2", "teal");
  styleMathLabel(SL.beta, SL.sigma, SL.gamma);

  const slRow = document.createElement("div");
  slRow.style.cssText = "display:flex;gap:20px;margin-bottom:12px;";
  slRow.appendChild(SL.beta.el);
  slRow.appendChild(SL.sigma.el);
  slRow.appendChild(SL.gamma.el);
  wrapper.appendChild(slRow);

  // ── Readout ──
  const readRow = document.createElement("div");
  readRow.style.cssText = "display:flex;gap:24px;align-items:baseline;margin-bottom:10px;";
  const rssBox = document.createElement("div");
  rssBox.style.cssText = "display:flex;flex-direction:column;gap:2px;";
  const rssLbl = document.createElement("span");
  rssLbl.style.cssText = "font-size:11px;font-weight:600;color:#64748b;text-transform:uppercase;letter-spacing:0.5px;";
  rssLbl.textContent = "∑ (yᵢ − ŷᵢ)² — RSS";
  const rssVal = document.createElement("span");
  rssVal.style.cssText = "font-size:26px;font-weight:800;color:#2563eb;font-family:'SF Mono',SFMono-Regular,Menlo,Consolas,monospace;";
  rssBox.appendChild(rssLbl);
  rssBox.appendChild(rssVal);
  const hoverBox = document.createElement("div");
  hoverBox.style.cssText = "font-size:13px;color:#475569;margin-left:auto;min-height:18px;";
  readRow.appendChild(rssBox);
  readRow.appendChild(hoverBox);
  wrapper.appendChild(readRow);

  // ── Plot ──
  // For the squares to look "square" in data units we need matching px/unit
  // on both axes. Work backwards from a target pixel scale.
  const unitPx = 14;                // 1 unit of incidence / 1 day = 14 px
  const innerW = unitPx * DAYS;
  const innerH = unitPx * 16;
  const M = { top: 14, right: 20, bottom: 34, left: 44 };
  const W = innerW + M.left + M.right;
  const H = innerH + M.top + M.bottom;

  const svg = d3.create("svg")
    .attr("viewBox", `0 0 ${W} ${H}`)
    .style("width", "100%").style("max-width", `${W}px`)
    .style("height", "auto").style("display", "block");

  const xSc = d3.scaleLinear().domain([0, DAYS]).range([0, innerW]);
  const ySc = d3.scaleLinear().domain([0, 16]).range([innerH, 0]);

  const g = svg.append("g").attr("transform", `translate(${M.left},${M.top})`);

  // Grid
  g.append("g").attr("stroke", "#e2e8f0").attr("stroke-dasharray", "2,3")
    .selectAll("line.yg").data(ySc.ticks(8)).join("line").attr("class","yg")
    .attr("x1", 0).attr("x2", innerW).attr("y1", d => ySc(d)).attr("y2", d => ySc(d));

  g.append("g").attr("transform", `translate(0,${innerH})`)
    .call(d3.axisBottom(xSc).ticks(7)).attr("font-size", 11);
  g.append("g").call(d3.axisLeft(ySc).ticks(8)).attr("font-size", 11);
  g.append("text")
    .attr("x", innerW/2).attr("y", innerH + 28)
    .attr("text-anchor", "middle")
    .style("font-size", "12px").style("fill", "#475569").text("Day");
  g.append("text")
    .attr("transform", `translate(-34,${innerH/2}) rotate(-90)`)
    .attr("text-anchor", "middle")
    .style("font-size", "12px").style("fill", "#475569").text("Incidence");

  // Squares layer (behind line)
  const gSq = g.append("g");
  // Line
  const linePath = g.append("path")
    .attr("fill", "none").attr("stroke", "#d97706").attr("stroke-width", 2.5);
  // Dots
  const dots = g.selectAll("circle.obs")
    .data(data).join("circle").attr("class","obs")
    .attr("cx", d => xSc(d.day)).attr("cy", d => ySc(d.inc))
    .attr("r", 3.2).attr("fill", "#0f172a");

  const lineGen = d3.line().x(d => xSc(d.t)).y(d => ySc(d.inc));

  // ── Buttons ──
  const btnRow = document.createElement("div");
  btnRow.style.cssText = "display:flex;gap:10px;margin-top:8px;max-width:320px;";
  const optBtn = createButton("▶  Optimize", "auto");
  const resetBtn = createButton("↻  Reset", "reset");
  btnRow.appendChild(optBtn.el);
  btnRow.appendChild(resetBtn.el);
  wrapper.appendChild(svg.node());
  wrapper.appendChild(btnRow);

  // ── Update ──
  let activeHover = null;

  function update() {
    const beta = SL.beta.val(), sigma = SL.sigma.val(), gamma = SL.gamma.val();
    const inc = mf_sim.run(beta, sigma, gamma, DAYS);
    linePath.attr("d", lineGen(inc.map((v, t) => ({ t, inc: v }))));

    const squares = [];
    let rss = 0;
    for (const d of data) {
      const p = inc[d.day];
      const r = d.inc - p;
      rss += r * r;
      const side = Math.abs(r);
      // Always draw square to the right of the day, baseline at the lower of (yObs, yPred)
      const yLow = Math.min(d.inc, p);
      squares.push({
        day: d.day, yObs: d.inc, yPred: p, resid: r, sqr: r*r,
        x0: d.day, x1: d.day + side,
        y0: yLow, y1: yLow + side
      });
    }

    rssVal.textContent = rss.toFixed(2);

    const sq = gSq.selectAll("rect.sq").data(squares, d => d.day);
    sq.join(
      enter => enter.append("rect").attr("class","sq")
        .attr("fill", "#3b82f6").attr("fill-opacity", 0.16)
        .attr("stroke", "#2563eb").attr("stroke-width", 1)
        .on("mouseenter", function(ev, d) {
          activeHover = d.day;
          d3.select(this).attr("fill-opacity", 0.45);
          hoverBox.innerHTML = `Day <b>${d.day}</b> &nbsp; residual = ${d.resid.toFixed(2)} &nbsp; residual² = <b>${d.sqr.toFixed(2)}</b>`;
        })
        .on("mouseleave", function() {
          activeHover = null;
          d3.select(this).attr("fill-opacity", 0.16);
          hoverBox.textContent = "";
        }),
      update => update,
      exit => exit.remove()
    )
      .attr("x", d => xSc(d.x0))
      .attr("y", d => ySc(d.y1))
      .attr("width",  d => Math.max(0, xSc(d.x1) - xSc(d.x0)))
      .attr("height", d => Math.max(0, ySc(d.y0) - ySc(d.y1)))
      .attr("fill-opacity", d => activeHover === d.day ? 0.45 : 0.16);

    // Residual stem (connects data point to its predicted value — always thin)
    const stem = gSq.selectAll("line.stem").data(squares, d => d.day);
    stem.join(
      enter => enter.append("line").attr("class","stem")
        .attr("stroke", "#2563eb").attr("stroke-width", 1.2),
      update => update,
      exit => exit.remove()
    )
      .attr("x1", d => xSc(d.day)).attr("x2", d => xSc(d.day))
      .attr("y1", d => ySc(d.yObs)).attr("y2", d => ySc(d.yPred));

    // Keep dots on top
    dots.raise();
  }

  SL.beta.input.addEventListener("input",  () => { SL.beta.sync();  update(); });
  SL.sigma.input.addEventListener("input", () => { SL.sigma.sync(); update(); });
  SL.gamma.input.addEventListener("input", () => { SL.gamma.sync(); update(); });

  function animateTo(target, ms = 600) {
    const start = { beta: SL.beta.val(), sigma: SL.sigma.val(), gamma: SL.gamma.val() };
    const t0 = performance.now();
    function frame(now) {
      const k = Math.min(1, (now - t0) / ms);
      const e = k < 0.5 ? 2*k*k : 1 - Math.pow(-2*k + 2, 2)/2;
      SL.beta.update( start.beta  + (target.beta  - start.beta)  * e);
      SL.sigma.update(start.sigma + (target.sigma - start.sigma) * e);
      SL.gamma.update(start.gamma + (target.gamma - start.gamma) * e);
      update();
      if (k < 1) requestAnimationFrame(frame);
    }
    requestAnimationFrame(frame);
  }

  optBtn.el.addEventListener("click", () =>
    animateTo({ beta: optim.beta, sigma: optim.sigma, gamma: optim.gamma }));
  resetBtn.el.addEventListener("click", () =>
    animateTo({ beta: 0.004, sigma: 0.5, gamma: 0.5 }));

  update();
  return wrapper;
}
```

$$RSS = \sum_{i = 1}^{n}[y_i - f(x_i)]^2$$ {#eq-calib-rss}

Least squares estimation corresponds to maximum likelihood estimation when the noise is normally distributed with equal variances.

### Maximum likelihood estimation {#sec-calib-mle}

Instead of minimising squared distance, **maximum likelihood** asks: *if our model were correct, how probable is the data we actually observed?* For count data like incidence, a natural choice is the **Poisson** distribution — each day's prediction $\hat{y}_i$ plays the role of the Poisson mean $\lambda_i$, and the probability of observing exactly $y_i$ cases is

$$P(Y_i = y_i \mid \lambda_i) = \frac{\lambda_i^{y_i}\, e^{-\lambda_i}}{y_i!}$$

The purple lollipops below show each day's Poisson PMF, with one stick per integer count. The **solid** stick sits at the observed $y_i$ — that height *is* its likelihood. Summing the log-likelihoods across days (and flipping sign) gives the negative log-likelihood, which optimisers minimise.

```{ojs}
//| echo: false

viewof mf_mle = {
  const data = transpose(ojsdata).filter(d => d.day <= 30);
  const DAYS = 30;

  const wrapper = document.createElement("div");
  wrapper.className = "mf-mle-wrap";
  wrapper.style.cssText = "font-family:system-ui,-apple-system,sans-serif;max-width:760px;margin:0 auto;";
  wrapper.appendChild(injectStyle());

  // One-time style sheet: styling via CSS classes so update() doesn't need per-element attr writes
  const styleTag = document.createElement("style");
  styleTag.textContent = `
    .mf-mle-wrap .stem { stroke:#7c3aed; stroke-width:1.4; stroke-opacity:0.38; }
    .mf-mle-wrap .stem.obs { stroke-width:2.6; stroke-opacity:0.95; }
    .mf-mle-wrap .head { fill:#a78bfa; fill-opacity:0.7; r:2; }
    .mf-mle-wrap .head.obs { fill:#7c3aed; fill-opacity:1; r:3.2; }
  `;
  wrapper.appendChild(styleTag);

  // Precomputed log-factorials (data counts are small — this covers any practical k)
  const LOG_FACT = new Float64Array(200);
  for (let i = 1; i < 200; i++) LOG_FACT[i] = LOG_FACT[i-1] + Math.log(i);
  const logPois = (k, lam) => {
    if (lam <= 0) return k === 0 ? 0 : -Infinity;
    return k * Math.log(lam) - lam - (k < LOG_FACT.length ? LOG_FACT[k] : LOG_FACT[LOG_FACT.length-1]);
  };

  // ── Sliders ──
  const SL = {};
  SL.beta  = createSlider("β",  0.001, 0.010, 0.0001, 0.004, "#3b82f6", "blue");
  SL.sigma = createSlider("σ",  0.3,   3.0,   0.05,   0.5,   "#7c3aed", "purple");
  SL.gamma = createSlider("γ",  0.3,   3.0,   0.05,   0.5,   "#0891b2", "teal");
  styleMathLabel(SL.beta, SL.sigma, SL.gamma);

  const slRow1 = document.createElement("div");
  slRow1.style.cssText = "display:flex;gap:20px;margin-bottom:12px;flex-wrap:wrap;";
  slRow1.appendChild(SL.beta.el);
  slRow1.appendChild(SL.sigma.el);
  slRow1.appendChild(SL.gamma.el);
  wrapper.appendChild(slRow1);

  // ── Readout ──
  const readRow = document.createElement("div");
  readRow.style.cssText = "display:flex;gap:24px;align-items:baseline;margin-bottom:10px;";
  const box = document.createElement("div");
  box.style.cssText = "display:flex;flex-direction:column;gap:2px;";
  const lbl = document.createElement("span");
  lbl.style.cssText = "font-size:11px;font-weight:600;color:#64748b;text-transform:uppercase;letter-spacing:0.5px;";
  lbl.textContent = "− log L(θ)  — Poisson";
  const valEl = document.createElement("span");
  valEl.style.cssText = "font-size:26px;font-weight:800;color:#7c3aed;font-family:'SF Mono',SFMono-Regular,Menlo,Consolas,monospace;";
  box.appendChild(lbl);
  box.appendChild(valEl);
  readRow.appendChild(box);
  wrapper.appendChild(readRow);

  // ── Plot ──
  const W = 720, H = 340;
  const M = { top: 14, right: 16, bottom: 34, left: 44 };
  const innerW = W - M.left - M.right;
  const innerH = H - M.top - M.bottom;

  const svg = d3.create("svg")
    .attr("viewBox", `0 0 ${W} ${H}`)
    .style("width", "100%").style("max-width", `${W}px`)
    .style("height", "auto").style("display", "block");

  const xSc = d3.scaleLinear().domain([0, DAYS]).range([0, innerW]);
  const ySc = d3.scaleLinear().domain([-1, 28]).range([innerH, 0]);

  const g = svg.append("g").attr("transform", `translate(${M.left},${M.top})`);

  g.append("g").attr("stroke", "#e2e8f0").attr("stroke-dasharray", "2,3")
    .selectAll("line.yg").data(ySc.ticks(6)).join("line").attr("class","yg")
    .attr("x1", 0).attr("x2", innerW).attr("y1", d => ySc(d)).attr("y2", d => ySc(d));

  g.append("g").attr("transform", `translate(0,${innerH})`)
    .call(d3.axisBottom(xSc).ticks(7)).attr("font-size", 11);
  g.append("g").call(d3.axisLeft(ySc).ticks(7)).attr("font-size", 11);
  g.append("text")
    .attr("x", innerW/2).attr("y", innerH + 28)
    .attr("text-anchor", "middle")
    .style("font-size", "12px").style("fill", "#475569").text("Day");
  g.append("text")
    .attr("transform", `translate(-34,${innerH/2}) rotate(-90)`)
    .attr("text-anchor", "middle")
    .style("font-size", "12px").style("fill", "#475569").text("Incidence");

  // Layers (draw order, bottom → top): fit line → stems → heads → observed dots
  const linePath = g.append("path")
    .attr("fill", "none").attr("stroke", "#d97706").attr("stroke-width", 2.5)
    .attr("stroke-opacity", 0.55);
  const gStems = g.append("g");
  const gHeads = g.append("g");
  const dots = g.selectAll("circle.obs")
    .data(data).join("circle").attr("class","obs")
    .attr("cx", d => xSc(d.day)).attr("cy", d => ySc(d.inc))
    .attr("r", 3.2).attr("fill", "#0f172a");

  const lineGen = d3.line().x(d => xSc(d.t)).y(d => ySc(d.inc));

  // ── Buttons ──
  const btnRow = document.createElement("div");
  btnRow.style.cssText = "display:flex;gap:10px;margin-top:8px;max-width:320px;";
  const optBtn = createButton("▶  Optimize", "auto");
  const resetBtn = createButton("↻  Reset", "reset");
  btnRow.appendChild(optBtn.el);
  btnRow.appendChild(resetBtn.el);
  wrapper.appendChild(svg.node());
  wrapper.appendChild(btnRow);

  const pxPerDay = xSc(1) - xSc(0);     // horizontal gap per day in SVG user units
  const maxStickPx = pxPerDay * 0.92;   // keep sticks inside the day's "cell"

  // Pre-allocate a fixed pool of line/circle elements per day; update() only mutates geometry.
  const MAX_K = 30;
  const stemGroups = [];
  const headGroups = [];

  for (let i = 0; i < data.length; i++) {
    const d = data[i];
    const stemArr = [];
    const headArr = [];
    for (let k = 0; k < MAX_K; k++) {
      const ln = gStems.append("line").attr("class", "stem").node();
      const ci = gHeads.append("circle").attr("class", "head").node();
      stemArr.push(ln);
      headArr.push(ci);
    }
    stemGroups.push(stemArr);
    headGroups.push(headArr);
  }

  function update() {
    const beta = SL.beta.val(), sigma = SL.sigma.val(), gamma = SL.gamma.val();
    const inc = mf_sim.run(beta, sigma, gamma, DAYS);
    linePath.attr("d", lineGen(inc.map((v, t) => ({ t, inc: v }))));

    let nll = 0;

    for (let i = 0; i < data.length; i++) {
      const d  = data[i];
      const mu = inc[d.day];
      const lli = logPois(d.inc, mu);
      nll += -lli;

      // Range of k around the mean to draw
      const sdApprox = Math.sqrt(Math.max(mu, 1));
      const kLo = Math.max(0, Math.floor(mu - 3.5 * sdApprox));
      const kHi = Math.min(MAX_K - 1, Math.ceil(mu + 3.5 * sdApprox) + 1);

      // Probability for each integer; compute peak for per-day scaling
      const kvals = new Array(MAX_K);
      let peak = 0;
      for (let k = kLo; k <= kHi; k++) {
        const p = Math.exp(logPois(k, mu));
        kvals[k] = p;
        if (p > peak) peak = p;
      }
      const scale = peak > 0 ? maxStickPx / peak : 0;
      const xBase = xSc(d.day);

      const stemArr = stemGroups[i];
      const headArr = headGroups[i];
      for (let k = 0; k < MAX_K; k++) {
        const ln = stemArr[k];
        const ci = headArr[k];
        if (k < kLo || k > kHi) {
          ln.setAttribute("visibility", "hidden");
          ci.setAttribute("visibility", "hidden");
          continue;
        }
        const p  = kvals[k] || 0;
        const xT = xBase + p * scale;
        const yP = ySc(k);
        ln.setAttribute("x1", xBase);
        ln.setAttribute("x2", xT);
        ln.setAttribute("y1", yP);
        ln.setAttribute("y2", yP);
        ln.removeAttribute("visibility");
        ci.setAttribute("cx", xT);
        ci.setAttribute("cy", yP);
        ci.removeAttribute("visibility");
        const isObs = (k === d.inc);
        if (isObs) { ln.classList.add("obs"); ci.classList.add("obs"); }
        else       { ln.classList.remove("obs"); ci.classList.remove("obs"); }
      }
    }

    valEl.textContent = nll.toFixed(2);
    dots.raise();
  }

  for (const k of ["beta","sigma","gamma"]) {
    SL[k].input.addEventListener("input", () => { SL[k].sync(); update(); });
  }

  function animateTo(target, ms = 600) {
    const start = {
      beta:  SL.beta.val(),
      sigma: SL.sigma.val(),
      gamma: SL.gamma.val()
    };
    const t0 = performance.now();
    function frame(now) {
      const k = Math.min(1, (now - t0) / ms);
      const e = k < 0.5 ? 2*k*k : 1 - Math.pow(-2*k + 2, 2)/2;
      SL.beta.update( start.beta  + (target.beta  - start.beta)  * e);
      SL.sigma.update(start.sigma + (target.sigma - start.sigma) * e);
      SL.gamma.update(start.gamma + (target.gamma - start.gamma) * e);
      update();
      if (k < 1) requestAnimationFrame(frame);
    }
    requestAnimationFrame(frame);
  }

  optBtn.el.addEventListener("click", () =>
    animateTo({ beta: optim_pois.beta, sigma: optim_pois.sigma, gamma: optim_pois.gamma }));
  resetBtn.el.addEventListener("click", () =>
    animateTo({ beta: 0.004, sigma: 0.5, gamma: 0.5 }));

  update();
  return wrapper;
}
```

A generalization of the least squares idea is the likelihood.

#### Likelihood

The likelihood is the probability of observing data $x$ given that our model is true[^2], $\mathbb{P}(x|M)$.

#### Log-likelihood {#sec-calib-loglik}

We have a lot of data points, and they are independent. The probability of observing all these data points at the same time is the production of these likelihood $\mathbb{P}(x_1|M) \times \mathbb{P}(x_2|M) \times \mathbb{P}(x_3|M)... = \prod\mathbb{P}(x|M)$.

Multiplying things together, you will end up losing precision if the numbers are too low. Here you are dealing with probability (a value \< 1), multiplying 100 probabilities you will end up with 1e-100.

But remember that $log(a \times b) = log(a) + log(b)$, and very convenient that $log(1^{-100} = -230.2585)$.

So $log(\mathbb{P}(x_1|M) \times \mathbb{P}(x_2|M) \times \mathbb{P}(x_3|M)...)$ $=$ $log(\mathbb{P}(x_1|M)) + log(\mathbb{P}(x_2|M)) + log(\mathbb{P}(x_3|M))...$ and it is so much easier to handle.

#### Negative log-likelihood {#sec-calib-negloglik}

Because statistical packages optimizers work by minimizing a function. Minimizing means **decrease the distance of two distributions** to its lowest, this is fairly easy because we get this when the distance close to 0 (just like the trick we do in hypothesis testing).

Minimizing negative log-likelihood (meaning that $-1 \times \text{log-likelihood}$) is equivalent to maximizing the log-likelihood, which is what we want to do (MLE: maximum likelihood estimation).

#### When MLE is equal to LSE {#sec-calib-mle-lse}

Assuming the data is generated from a normal distribution with mean $\mu$ and a standard deviation $\sigma$.

-   **Step 1**. Write down the likelihood function $L(\theta)$.

$$L(\theta) = \prod_{i = 1}^{n} \mathcal{N}(\mu_i, \sigma) = \prod_{i = 1}^{n} \frac{1}{\sigma\sqrt{2 \pi}} \text{exp} \left[ \frac{-1}{2 \sigma^2} (y_i - \mu_i)^2 \right]$$

$$L(\theta) = \frac{1}{(\sigma \sqrt{2 \pi})^n} \text{exp} \left[ \frac{-1}{2 \sigma^2} \sum_{i = 1}^n (y_i - \mu_i)^2 \right]$$

Since $\sigma$ and $\sqrt{2 \pi}$ are constant.

$$L(\theta) \propto \text{exp} \left[ - \sum_{i = 1}^n (y_i - \mu_i)^2 \right]$$

-   **Step 2**. Take the natural log of the likelihood $log L(\theta)$.

$$log L(\theta) \propto - \sum_{i = 1}^n (y_i - \mu_i)^2$$

-   **Step 3**. Take the negative log-likelihood $- log L(\theta)$.

$$- log L(\theta) \propto \sum_{i = 1}^n (y_i - \mu_i)^2$$

This looks exactly like the residual sum of squares at @eq-calib-rss.

-   **Step 4**. The optimizer **minimize negative log-likelihood** $- log L(\theta)$. It does the same thing as the LSE finds the optimal parameters by **minimizing the RSS**.

Nelder-Mead is a numerical algorithm to find the minimum (or maximum) of a multidimentional function. It's a direct search method: calculate the value of the function and compare it to other values. It does not use derivatives.

It use a ***simplex*** for a search. A simplex is a

## Uncertainty

Uncertainties in the parameter estimates are described by confidence intervals [@bolker2008]. A confidence interval $[\sigma_i^-, \sigma_i^+]$ of a parameter estimate $\hat{\theta_i}$ to a confidence level $1 - \alpha$ signifies that the true value $\theta_i^*$ is located within this interval with probability $1 - \alpha$ [@bolker2008].
