# Estimating $R_0$

## Exponential growth rate

Step 1. Find an estimate for the mean generation time $T_g$.

Step 2. Identify the exponential phase of the epidemic curve: the period of which $\log(\text{incidence})$ is linear.

Step 3. Estimate the growth rate $r$ for that phase by finding the slope from the linear regression on the $\log(\text{incidence})$.

::: {.callout-tip collapse="true"}

## Proof

See @sec-rnum-growthrate-r0 for more details. During the exponential phase, the incidence at time $t$ is given by:

$$C_t = C_0 e^{rt}$$

Take the natural logarithm:

$$\log C_t = log C_0 + rt$$

Therefore $r$ is the slope from the linear regression on the $\log(\text{incidence})$.

:::

Step 4. Given the exponential growth rate $r$ and the $T_g$, calculate the basic reproduction number:

- In an SIR model $T_g = \frac{1}{\gamma}$ so:

$$R_0 \approx 1 + \frac{r}{\gamma}$$

- In an SEIR model $T_g = \frac{1}{\gamma} + \frac{1}{\sigma}$ so:

$$R_0 \approx \left( 1 + \frac{r}{\gamma} \right) \left( 1 + \frac{r}{\sigma} \right)$$

### Code

```{r, message=FALSE, warning=FALSE}
library(R0)
library(ggplot2)
```

```{r}
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "100%"
df <- read.csv("data/flu_inc.csv")

ggplot(df, aes(x = day, y = inc)) +
  geom_point() +
  labs(x = "Day", y = "Incidence") +
  theme_minimal()
```

```{r}
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "100%"
df$loginc <- log(df$inc)

ggplot(df, aes(x = day, y = loginc)) +
  geom_point() +
  labs(x = "Day", y = "Log incidence") +
  theme_minimal()

lm(loginc ~ day, data = df[5:13,])
```

### 

```{r}
1 + 0.09*4

gt <- generation.time("gamma", c(3, 1.5))
tmp <- data.frame()
est.R0.EG(df$inc, gt, begin = 5, end = 13)
```

## Simple compartmental models

Consider a simple closed population SIR model.

$$\begin{align} \frac{dS}{dt} & = -\beta SI \\ \frac{dI}{dt} & = \beta SI - \gamma I\\ \frac{dR}{dt} & = \gamma I \end{align}$$

Focusing on the $I$ equation:

$$\frac{dI}{dt} = \beta SI - \gamma I = I (\beta S - \gamma)$$

A disease spreads when:

$$\begin{align} \frac{dI}{dt} & > 0 \\ \Leftrightarrow I (\beta S - \gamma) & > 0 \\ \Leftrightarrow \beta S - \gamma & > 0 \\ \Leftrightarrow \beta S & > \gamma \\ \Leftrightarrow \frac{\beta S}{\gamma} & > 1 \end{align}$$

This leads to the definition of $R_0$. By @def-r0, $R_0$ is the $R$ in a fully susceptible population $S_0$:

$$R_0 = \frac{\beta S_0}{\gamma}$$

Therefore:

-   $R_0 > 1$ meaning that $\frac{dI}{dt} > 0$, then the disease will spread.
-   $R_0 < 1$ meaning that $\frac{dI}{dt} < 0$, then no epidemic occurs.

For more complex compartmental models, especially those with more infected compartments, as proposed by [@diekmann1990] and [@vandendriessche2002], $R_0$ can be determined using the next generation matrix [@vandendriessche2017].

## Next generation matrix

$R_0$ is the maximum absolute eigenvalue (also called the spectral radius) of the next generation matrix (NGM) [@vandendriessche2017].

Consider a SEIR model with equal birth and death rate $\mu$.

$$\begin{align} \frac{dS}{dt} & = \mu (S + E + I + R) -\beta SI - \mu S \\ \frac{dE}{dt} & = \beta SI - \sigma E - \mu E \\ \frac{dI}{dt} & = \sigma E - \gamma I - \mu I \\ \frac{dR}{dt} & = \gamma I - \mu R \end{align}$$

**Step 1. Identify the infected compartments**

For our SEIR model, the infected compartments are $E$ and $I$.

**Step 2. Write the NGM functions**

-   $\mathcal{F}_i(x)$: is the rate of appearance of new infections in compartment $i$ [@vandendriessche2017] (caused by contact).
-   $\mathcal{V}_i(x)$: is the rate of other transitions between compartment $i$ and other infected compartments [@vandendriessche2017] (caused by transition between compartments).

For our SEIR model, we would write $\mathcal{F}$ and $\mathcal{V}$ as follows:

::: columns
::: {.column width="50%"}
$$\mathcal{F} = \begin{bmatrix} \beta SI \\ 0 \end{bmatrix}$$ $\mathcal{F}$ is a vector contains of $\begin{bmatrix} \mathcal{F}_1 \\ \mathcal{F}_2 \end{bmatrix}$:

-   $\mathcal{F}_1 = \beta SI$: the rate of new infections (caused by contact) in compartment $E$.
-   $\mathcal{F}_2 = 0$: as there is no new infections (caused by contact) in compartment $I$.
:::

::: {.column width="50%"}
$$\mathcal{V} = \begin{bmatrix} \sigma E + \mu E \\ - \sigma E + \gamma I + \mu I \end{bmatrix}$$

$\mathcal{V}$ is a vector contains of $\begin{bmatrix} \mathcal{V}_1 \\ \mathcal{V}_2 \end{bmatrix}$:

-   $\mathcal{V}_1 = \sigma E + \mu E$: is the rate of other transitions to and from compartment $E$.
-   $\mathcal{V}_2 = - \sigma E + \gamma I + \mu I$: is the rate of other transitions to and from compartment $I$.
:::
:::

$\mathcal{F}$ and $\mathcal{V}$ are written such that:

$$\frac{dx}{dt} = \mathcal{F} - \mathcal{V}$$

Here we have:

$$\frac{dE}{dt} = \mathcal{F}_1 - \mathcal{V}_1 = \beta SI - (\sigma E + \mu E)$$

$$\frac{dI}{dt} = \mathcal{F}_2 - \mathcal{V}_2 = 0 - (- \sigma E + \gamma I + \mu I)$$

**Step 3. Find the Jacobian matrix**

Jacobian matrix $F$ for $\mathcal{F}$, and $V$ for $\mathcal{V}$. Jacobian matrix is a matrix contain all partial derivatives of a vector function.

$$\begin{align} F & = \begin{bmatrix} \frac{\partial \mathcal{F}_1}{\partial E} & \frac{\partial \mathcal{F}_1}{\partial I} \\ \frac{\partial \mathcal{F}_2}{\partial E} & \frac{\partial \mathcal{F}_2}{\partial I} \end{bmatrix} \\ & = \begin{bmatrix} \frac{\partial [\beta SI]}{\partial E} & \frac{\partial [\beta SI]}{\partial I} \\ \frac{\partial [0]}{\partial E} & \frac{\partial [0]}{\partial I} \end{bmatrix} \\ & = \begin{bmatrix} 0 & \beta S \\ 0 & 0 \end{bmatrix} \end{align}$$

In which:

-   $\frac{\partial [\beta SI]}{\partial E} = 0$: because there is no $E$ in $\beta SI$.
-   $\frac{\partial [\beta SI]}{\partial I} = \beta S$.

$$\begin{align} V & = \begin{bmatrix} \frac{\partial \mathcal{V}_1}{\partial E} & \frac{\partial \mathcal{V}_1}{\partial I} \\ \frac{\partial \mathcal{V}_2}{\partial E} & \frac{\partial \mathcal{V}_2}{\partial I} \end{bmatrix} \\ & = \begin{bmatrix} \frac{\partial [\sigma E + \mu E]}{\partial E} & \frac{\partial [\sigma E + \mu E]}{\partial I} \\ \frac{\partial [- \sigma E + \gamma I + \mu I]}{\partial E} & \frac{\partial [- \sigma E + \gamma I + \mu I]}{\partial I} \end{bmatrix} \\ & = \begin{bmatrix} \sigma + \mu & 0 \\ - \sigma & \gamma + \mu \end{bmatrix} \end{align}$$

**Step 4. Compute the NGM**

Use @def-inv-matrix to compute inverse matrix $V^{-1}$:

$$\begin{align} V^{-1} & = \frac{1}{(\sigma + \mu)(\gamma + \mu)} \begin{bmatrix} \gamma + \mu & - 0 \\ - \sigma & \sigma + \mu \end{bmatrix} \\& = \begin{bmatrix} \frac{1}{\sigma + \mu} & 0 \\ \frac{\sigma}{(\sigma + \mu)(\gamma + \mu)} & \frac{1}{\gamma + \mu} \end{bmatrix} \end{align}$$

$$\begin{align} K & = FV^{-1} \\ & = \begin{bmatrix} 0 & \beta S \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \frac{1}{\sigma + \mu} & 0 \\ \frac{\sigma}{(\sigma + \mu)(\gamma + \mu)} & \frac{1}{\gamma + \mu} \end{bmatrix} \\ & = \begin{bmatrix} \frac{\beta S \sigma}{(\sigma + \mu)(\gamma + \mu)} & \frac{\beta S}{\gamma + \mu} \\ 0 & 0 \end{bmatrix} \end{align}$$

::: {.callout-tip collapse="true"}
## What is a next generation matrix?

Source: quantpie's [video](https://www.youtube.com/watch?v=xJ5mizenTII).

A population consists of two groups: young and old.

-   A young person on average infects 3 young and 2 old: $\begin{bmatrix} 3 \\ 2 \end{bmatrix}$.
-   An old person on average infects 1 young and 2 old: $\begin{bmatrix} 1 \\ 2 \end{bmatrix}$.

Each group becomes a column in the next generation matrix (NGM):

$$\text{NGM} = \begin{bmatrix} 3 & 1 \\ 2 & 2 \end{bmatrix}$$

$R_0$ is the root of equation:

$$\begin{align} \begin{bmatrix} 3 - \lambda & 1 \\ 2 & 2 - \lambda \end{bmatrix} & = 0 \\ \Leftrightarrow (3 - \lambda) \times (2 - \lambda) - 2 \times 1 & = 0 \\ \Leftrightarrow \lambda^2 - 5 \lambda + 6 - 2 & = 0 \\ \Leftrightarrow \lambda^2 - 5 \lambda + 4 & = 0 \end{align}$$

There are two roots $\lambda = 4$ and $\lambda = 1$, for an outbreak to occur $R_0 > 1$, so here $R_0 = 4$.
:::

**Step 5. Compute** $R_0$

$R_0$ is the eigenvalues of the NGM.

$$\begin{align} det(K - \lambda) & = \begin{bmatrix} \frac{\beta S \sigma}{(\sigma + \mu)(\gamma + \mu)} - \lambda & \frac{\beta S}{\gamma + \mu} \\ 0 & 0 - \lambda \end{bmatrix} \\ & = [\frac{\beta S \sigma}{(\sigma + \mu)(\gamma + \mu)} - \lambda][0 - \lambda] \end{align}$$

$$R_0 = \frac{\beta S \sigma}{(\sigma + \mu)(\gamma + \mu)}$$

::: {.callout-note collapse="true"}
## Use NGM to compute $R_0$ for SIR model

Consider a simple closed population SIR model.

$$\begin{align} \frac{dS}{dt} & = -\beta SI \\ \frac{dI}{dt} & = \beta SI - \gamma I\\ \frac{dR}{dt} & = \gamma I \end{align}$$

**Step 1. Identify the infected compartments**

For our SIR model, the infected compartment is $I$.

**Step 2. Write the NGM functions**

$$\mathcal{F} = \beta SI$$

$$\mathcal{V} = \gamma I$$

Such that we have:

$$\frac{dI}{dt} = \mathcal{F} - \mathcal{V} = \beta SI - \gamma I$$

**Step 3. Find the Jacobian matrix**

$$F = \frac{\partial \mathcal{F}}{\partial I} = \frac{\partial [\beta SI]}{\partial I} = \beta S$$

$$V = \frac{\partial \mathcal{V}}{\partial I} = \frac{\partial [\gamma I]}{\partial I} = \gamma$$

**Step 4 and 5. Compute the NGM and** $R_0$

$$V^{-1} = \gamma^{-1} = \frac{1}{\gamma}$$

$$K = FV^{-1} = \frac{\beta S}{\gamma}$$

Here we also have $R_0$.

$$R_0 = \frac{\beta S}{\gamma}$$
:::

## Epidemic final size

If there were no interventions or changes in behavior, the proportion of the population infected during the entire course of an epidemic would be approximately the non-zero solution of the equation:

\$\$\$\$
