# SIR

## Assumptions

-   Each member of a population is either susceptible, infectious, or recovered [@vandendriessche2017].
-   The disease spreads through direct contact between susceptible and infected individuals.
-   Well-mixed population, or homogeneous mixing (represented by a single contact rate $\beta$) [@becker2017].
-   In the most basic form it considers a "closed population" without demographics (no births, deaths, or migration) [@keeling2011], make sense when the disease is short lived compared with the population lifetime [@vandendriessche2017].
-   Infectious disease properties are for a single pathogen [@becker2017].
-   Recovered individuals gain lifelong immunity.

## Model

$$\frac{dS}{dt} = -\beta SI = -\lambda S$$ {#eq-sir-ds}

$$\frac{dI}{dt} = \beta SI - \gamma I = \lambda S - \gamma I$$ {#eq-sir-di}

$$\frac{dR}{dt} = \gamma I$$ {#eq-sir-dr}

-   $\beta$: effective contact rate.
-   $\gamma$: removal rate.
-   $\beta I = \lambda$: the rate from S to I is called **force of infection**.

## Code

```{r, message=FALSE, warning=FALSE}
library(deSolve)
library(tidyr)
library(bbmle)
library(ggplot2)
library(RColorBrewer)
library(patchwork)
```

Below is how we solve the differential equations @eq-sir-ds, @eq-sir-di and @eq-sir-dr in R with the `deSolve` package.

::: {#lst-sir-det}

```{r}
sir_mod <- function(beta, gamma, S0, I0, R0, times) {
  
  equations <- function(time, variables, parameters) {
    with(as.list(c(variables, parameters)), {
      dS <-   -beta * S * I
      dI <-    beta * S * I - gamma * I
      dR <-    gamma * I
      dCInc <- beta * S * I
      return(list(c(dS, dI, dR, dCInc)))
    })
  }
  
  parameters_values <- c(beta = beta, gamma = gamma)
  initial_values <- c(S = S0, I = I0, R = R0, CInc = 0)
  
  out <- ode(y = initial_values, times = times, func = equations, parms = parameters_values)

  out <- as.data.frame(out)
  # Compute incidence from cumulative incidence
  out$Inc <- c(I0, diff(out$CInc))
  out$CInc <- NULL
  out
}
```

Deterministic SIR model

:::

::: {.callout-important}
Be careful of what kind of data we are using to fit the model:

-   **Incidence**: use the $\beta SI$ part.

$$\text{Incidence} = \frac{\text{Number of new cases during a period}}{\text{Population at risk during the same period}}$$

-   **Prevalence**: use the **I compartment**.

$$\text{Prevalence} = \frac{\text{Number of existing cases at time t}}{\text{Total population at time t}}$$
:::

There are two ways to compute incidence.

1. While writing the differential equations, we tell `deSolve` to compute the cumulative incidence, then compute the incidence by using `diff()` on the cumulative incidence.

```{r, eval=F}
# Cumulative incidence
dCInc <- beta * S * I
# Compute incidence from cumulative incidence
out$Inc <- c(I0, diff(out$CInc))
```

2. In this "closed population", the incidence is the number of individuals moving out of $S$ in each time step. So we can compute it from the $S$ column. Let compare it with the previous method.

```{r}
pred <- sir_mod(beta = 0.004, gamma = 0.5, S0 = 999, I0 = 1, R0 = 0, times = 0:13)

tmp <- pred
tmp$Inc2 <- c(1, -diff(tmp$S))
tmp[,"Inc-Inc2"] <- round(tmp$Inc - tmp$Inc2)
tmp[,c("Inc", "Inc2", "Inc-Inc2")]
```

Now let have a look at the output.

```{r}
head(pred)
```

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "100%"
df_plot <- pivot_longer(pred, cols = S:Inc, names_to = "comp", values_to = "n")

ggplot(df_plot, aes(x = time, y = n, color = comp)) +
  geom_point() + geom_line() +
  scale_color_brewer(palette = "PuOr", breaks = c("S", "Inc", "I", "R")) +
  labs(color = NULL, y = NULL, x = "Time") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The time to peak of incidence is earlier than prevalence:

- Incidence reflects the immediate rate of new infections, the peak of incidence is reached when the transmission rate begins to decline.
- Prevalence accumulates both new and existing cases. Even after the incidence starts to decline, the active cases who were infected before the incidence peak are still there waiting for recovering. Prevalence only starts to decline when the recovery rate surpasses the rate of new infections significantly.

## Sensitivity analysis

The effective contact rate $\beta$ and the recovery rate $\gamma$ both affect whether or not there is an epidemic. The total number of people in the population and the initial number of infected people have no effect on whether there is an epidemic.

Below is a function that take a range of values of parameters and compare the model outputs.

```{r}
#| code-fold: true
sir_sa <- function(vals, parm, outcome, times, init_beta = 0.004, init_gamma = 0.5) {
  pick_comp <- ifelse(outcome == "prev", "I", "Inc")
  plot_lab <- ifelse(outcome == "prev", "Prevalence", "Incidence")
  symb <- ifelse(parm == "beta", "\u03b2", "\u03b3")
  l <- list()
  for(i in 1:length(vals)) {
    if (parm == "beta") {
      pred <- sir_mod(beta = vals[i], gamma = init_gamma, S0 = 999, I0 = 1, R0 = 0, times = times)[, c("time", pick_comp)]
    } else {
      pred <- sir_mod(beta = init_beta, gamma = vals[i], S0 = 999, I0 = 1, R0 = 0, times = times)[, c("time", pick_comp)]
    }
    pred$comp <- paste0(pick_comp, i)
    colnames(pred) <- c("time", "n", "comp")
    l[[i]] <- pred
  }
  # Bind this list into a data frame
  df_plot <- Reduce(rbind, l)
  # Set level I1 < I2 < I3... according to the increase of beta values
  df_plot$comp <- factor(df_plot$comp, levels = c(paste0(pick_comp, 1:length(vals))))
  
  # Plot results
  my_palette <- brewer.pal(9, "Blues")[(9-length(vals)+1):9]
  
  ggplot(df_plot, aes(x = time, y = n, color = comp)) + 
    geom_point() + geom_line() +
    scale_color_manual(values = my_palette, labels = c(paste0(symb, " = ", vals))) +
    labs(color = NULL, y = plot_lab, x = "Time") +
    theme_minimal()
}
```

### Effective contact rate $\beta$

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 4
#| out-width: "100%"

vals <- seq(0.001, 0.006, 0.001)

p1 <- sir_sa(vals = vals, parm = "beta", outcome = "prev", times = 0:13) + labs(x = NULL)
p2 <- sir_sa(vals = vals, parm = "beta", outcome = "inc", times = 0:13)

p1 / p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

When $\beta$ **increases**, more susceptible individuals become infected per unit time:

- Outbreak size **increases**: $R_0 = \frac{\beta}{\gamma}$, so if $\beta$ increases, $R_0$ **increases**, each infected individual infect more people on average.
- **Larger** epidemic peak: the peak number at any given time will be higher, because more susceptible individuals are infected.
- Time until peak ***decreases***: because the speed of susceptible individuals become infected is faster.
- Epidemic duration ***decreases***: because the pool of susceptible individuals depletes quickly, the epidemic may conclude sooner when there are few susceptible individuals left.

### Recovery rate $\gamma$

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 4
#| out-width: "100%"

vals <- seq(0.3, 0.8, 0.1)

p1 <- sir_sa(vals = vals, parm = "gamma", outcome = "prev", times = 0:13) + labs(x = NULL)
p2 <- sir_sa(vals = vals, parm = "gamma", outcome = "inc", times = 0:13)

p1 / p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

When $\gamma$ **increases**, patients recover with a faster speed:

- Outbreak size ***decreases***: patients have ***less time to infect*** susceptible individuals; and also $R_0 = \frac{\beta}{\gamma}$, so if $\gamma$ increases, $R_0$ ***decreases***, each infected individual infect less people on average.
- ***Smaller*** epidemic peak: the peak number of infected individuals will be lower because $R_0$ ***decreases***.
- Time until peak ***slightly decreases***: because the infected population clears out faster, but this depends on the balance between the faster recovery and the rate of new infections.
- Epidemic duration ***slightly decreases***: for same reason as the peak time.

## Corollaries

### The threshold phenomenon

From @eq-sir-di:

$$\frac{dI}{dt} = \beta SI - \gamma I = I(\beta S - \gamma)$$

The infection "dies out" when:

$$\frac{dI}{dt} < 0 \Leftrightarrow \beta S - \gamma < 0 \Leftrightarrow S < \frac{\gamma}{\beta}$$

### Epidemic burnout

Divide @eq-sir-ds by @eq-sir-dr:

$$\frac{dS}{dR} = \frac{-\beta S I}{\gamma I} = -\frac{\beta}{\gamma} S$$

Upon integrating with respect to R, we obtain:

$$S(t) = S(0)e^{-\frac{\beta}{\gamma} R(t)}$$

Given that $e^x > 0 \quad \forall x$, therefore $S(t) > 0$ there will always be some susceptibles in the population who escape infection, leading to an important counter-intuitive conclusion from this model:

> The chain of transmission eventually breaks due to the decline in infectives, not due to a complete lack of susceptibles.
