---
title: "SIR"
---

```{ojs}
//| echo: false
import { createSlider, createButton, injectStyle, styleMathLabel } from "./_slider.js"
d3 = require("d3@7")
```


## Model {#sec-sir-mod}

![](img/det/sir.svg)

$$\frac{dS}{dt} = -\beta SI = -\lambda S$$ {#eq-sir-ds}

$$\frac{dI}{dt} = \beta SI - \gamma I = \lambda S - \gamma I$$ {#eq-sir-di}

$$\frac{dR}{dt} = \gamma I$$ {#eq-sir-dr}

-   $\beta$: effective contact rate.
-   $\gamma$: removal rate.
-   $\beta I = \lambda$: the rate from S to I is called **force of infection**.

## Assumptions

-   Each member of a population is either susceptible, infectious, or recovered [@vandendriessche2017].
-   The disease spreads through direct contact between susceptible and infected individuals.
-   Well-mixed population, or homogeneous mixing (represented by a single effective contact rate $\beta$) [@becker2017].
-   In the most basic form it considers a "closed population" without demographics (no births, deaths, or migration) [@keeling2011], make sense when the disease is short lived compared with the population lifetime [@vandendriessche2017].
-   There is no latent period, individuals become infectious immediately upon contracting the disease.
-   The effective contact rate $\beta$ and recovery rate $\gamma$ are constant over time.
-   Infectious disease properties are for a single pathogen [@becker2017].
-   Recovered individuals gain lifelong immunity.

```{ojs}
//| echo: false

viewof sir_sim = (() => {
  // ══════════════════════════════════════════════════════
  // 1. WRAPPER + SHARED STYLE FROM _slider.js
  // ══════════════════════════════════════════════════════
  const wrapper = document.createElement("div");
  wrapper.style.cssText = `
    display:flex;flex-direction:column;align-items:center;
    width:100%;max-width:960px;margin:0 auto;
  `;

  wrapper.appendChild(injectStyle());

  // ══════════════════════════════════════════════════════
  // 2. SLIDERS  (initial S, I, R + β, γ)
  // ══════════════════════════════════════════════════════
  const SL = {};
  SL.s0    = createSlider("Initial Susceptible", 50, 400, 10, 200, "#3b82f6", "blue");
  SL.i0    = createSlider("Initial Infected",     1,  50,  1,   5, "#dc2626", "red");
  SL.r0    = createSlider("Initial Recovered",    0, 100,  5,   0, "#16a34a", "green");
  SL.beta  = createSlider("β", 0.05, 1.00, 0.05, 0.40, "#d97706", "amber");
  SL.gamma = createSlider("γ", 0.05, 0.50, 0.05, 0.10, "#7c3aed", "purple");
  styleMathLabel(SL.beta, SL.gamma);

  const r1 = document.createElement("div");
  r1.style.cssText = "display:flex;gap:24px;width:100%;margin-bottom:10px;";
  r1.appendChild(SL.s0.el); r1.appendChild(SL.i0.el); r1.appendChild(SL.r0.el);
  wrapper.appendChild(r1);

  const r2 = document.createElement("div");
  r2.style.cssText = "display:flex;gap:24px;width:100%;margin-bottom:12px;";
  r2.appendChild(SL.beta.el); r2.appendChild(SL.gamma.el);
  wrapper.appendChild(r2);

  const r3 = document.createElement("div");
  r3.style.cssText = "display:flex;gap:10px;width:100%;margin-bottom:14px;padding-bottom:14px;border-bottom:1px solid #e2e8f0;";
  const btnStep  = createButton("▶  Step",  "step");
  const btnAuto  = createButton("⏩  Auto",  "auto");
  const btnReset = createButton("↺  Reset", "reset");
  r3.appendChild(btnStep.el); r3.appendChild(btnAuto.el); r3.appendChild(btnReset.el);
  wrapper.appendChild(r3);

  // ══════════════════════════════════════════════════════
  // 3. STATS LINE  (Day · R₀ · N)
  // ══════════════════════════════════════════════════════
  const stats = document.createElement("div");
  stats.style.cssText = `display:flex;justify-content:space-between;width:100%;font-size:13px;color:#334155;margin:0 4px 12px;font-family:"SF Mono",SFMono-Regular,Menlo,Consolas,monospace;`;
  const dayLabel = document.createElement("span"); dayLabel.style.fontWeight = "700";
  const r0Label  = document.createElement("span");
  const popLabel = document.createElement("span");
  stats.appendChild(dayLabel); stats.appendChild(r0Label); stats.appendChild(popLabel);
  wrapper.appendChild(stats);

  // ══════════════════════════════════════════════════════
  // 4. SVG HELPERS
  // ══════════════════════════════════════════════════════
  const NS = "http://www.w3.org/2000/svg";
  function el(tag, a) {
    const e = document.createElementNS(NS, tag);
    if (a) for (const [k, v] of Object.entries(a)) e.setAttribute(k, v);
    return e;
  }

  // ══════════════════════════════════════════════════════
  // 5. COMPARTMENT SVG  (S · I · R boxes — one dot group per box)
  // ══════════════════════════════════════════════════════
  const W = 760, H = 300, BOX_GAP = 12;
  const boxW = Math.floor((W - 2 * BOX_GAP) / 3);
  const boxY = 6, boxH = H - 2 * boxY;

  const compartments = [
    { key: "S", label: "SUSCEPTIBLE", color: "#3b82f6", border: "#bfdbfe", bg: "#eff6ff", x: 0 },
    { key: "I", label: "INFECTED",    color: "#dc2626", border: "#fecaca", bg: "#fef2f2", x: boxW + BOX_GAP },
    { key: "R", label: "RECOVERED",   color: "#16a34a", border: "#bbf7d0", bg: "#f0fdf4", x: 2 * (boxW + BOX_GAP) },
  ];

  const compSvg = el("svg");
  compSvg.setAttribute("viewBox", `0 0 ${W} ${H}`);
  compSvg.style.cssText = `width:100%;max-width:${W}px;margin-bottom:14px;`;

  const counts = {}, overflow = {}, dotG = {};
  for (const c of compartments) {
    compSvg.appendChild(el("rect", { x: c.x, y: boxY, width: boxW, height: boxH, rx: 12, fill: c.bg, stroke: c.border, "stroke-width": 2 }));

    const lbl = el("text", { x: c.x + 14, y: 26, fill: c.color, "font-size": 12, "font-weight": 700, "font-family": "system-ui, -apple-system, sans-serif" });
    lbl.style.letterSpacing = "0.6px"; lbl.textContent = c.label;
    compSvg.appendChild(lbl);

    const cnt = el("text", { x: c.x + boxW - 14, y: 32, fill: c.color, "font-size": 24, "font-weight": 800, "font-family": "'SF Mono',monospace", "text-anchor": "end" });
    cnt.textContent = "0"; counts[c.key] = cnt;
    compSvg.appendChild(cnt);

    compSvg.appendChild(el("line", { x1: c.x + 10, x2: c.x + boxW - 10, y1: 42, y2: 42, stroke: c.border }));

    const ov = el("text", { x: c.x + boxW - 10, y: H - 10, fill: c.color, "font-size": 10, "font-family": "'SF Mono',monospace", "text-anchor": "end", opacity: 0.65 });
    overflow[c.key] = ov;
    compSvg.appendChild(ov);

    // Per-compartment dot group — dots only ever belong to one box, no flying across.
    const g = el("g");
    dotG[c.key] = g;
    compSvg.appendChild(g);
  }
  wrapper.appendChild(compSvg);

  // ══════════════════════════════════════════════════════
  // 6. TIME-SERIES SVG  (S, I, R lines vs day)
  // ══════════════════════════════════════════════════════
  const TW = 760, TH = 280;
  const tmg = { t: 16, r: 96, b: 40, l: 56 };
  const tcw = TW - tmg.l - tmg.r, tch = TH - tmg.t - tmg.b;

  const tsSvg = el("svg");
  tsSvg.setAttribute("viewBox", `0 0 ${TW} ${TH}`);
  tsSvg.style.cssText = `width:100%;max-width:${TW}px;background:#fafbfc;border-radius:10px;border:1px solid #e2e8f0;`;

  const gridLines = [], yLabels = [];
  for (let i = 0; i <= 5; i++) {
    gridLines.push(tsSvg.appendChild(el("line", { stroke: "#e2e8f0", "stroke-width": 0.7 })));
    yLabels.push(tsSvg.appendChild(el("text", { "text-anchor": "end", fill: "#94a3b8", "font-size": 10, "font-family": "'SF Mono',monospace" })));
  }
  const xAx = tsSvg.appendChild(el("line", { stroke: "#94a3b8" }));
  const yAx = tsSvg.appendChild(el("line", { stroke: "#94a3b8" }));

  const xTickPool = [];
  for (let i = 0; i < 24; i++) {
    const ln = tsSvg.appendChild(el("line", { stroke: "#94a3b8", y2: 5 }));
    const t  = tsSvg.appendChild(el("text", { "text-anchor": "middle", fill: "#64748b", "font-size": 11, "font-family": "'SF Mono',monospace", dy: 16 }));
    ln.style.display = "none"; t.style.display = "none";
    xTickPool.push({ ln, t });
  }
  const xLbl = tsSvg.appendChild(el("text", { "text-anchor": "middle", fill: "#64748b", "font-size": 12, y: TH - 6 }));
  xLbl.textContent = "Day";
  const yLbl = tsSvg.appendChild(el("text", { "text-anchor": "middle", fill: "#64748b", "font-size": 12, x: 14, y: tmg.t + tch / 2 }));
  yLbl.setAttribute("transform", `rotate(-90,14,${tmg.t + tch / 2})`);
  yLbl.textContent = "Number of individuals";

  // Draw R first, then S, then I on top (I is the line viewers track most)
  const pathR_ = tsSvg.appendChild(el("path", { fill: "none", stroke: "#16a34a", "stroke-width": 2.5, "stroke-linejoin": "round", "stroke-linecap": "round" }));
  const pathS_ = tsSvg.appendChild(el("path", { fill: "none", stroke: "#3b82f6", "stroke-width": 2.5, "stroke-linejoin": "round", "stroke-linecap": "round" }));
  const pathI_ = tsSvg.appendChild(el("path", { fill: "none", stroke: "#dc2626", "stroke-width": 2.5, "stroke-linejoin": "round", "stroke-linecap": "round" }));

  // "Now" reference line  (subtle dashed vertical at current day)
  const nowLine = tsSvg.appendChild(el("line", { stroke: "#cbd5e1", "stroke-dasharray": "3 3", "stroke-width": 1 }));
  nowLine.style.display = "none";

  // End-of-line labels stamped on the right edge of the visible data
  const endDots = {}, endTexts = {};
  ['S', 'I', 'R'].forEach(k => {
    const color = { S: '#3b82f6', I: '#dc2626', R: '#16a34a' }[k];
    const dot = tsSvg.appendChild(el("circle", { r: 3.5, fill: color }));
    dot.style.display = "none"; endDots[k] = dot;
    const t = tsSvg.appendChild(el("text", { fill: color, "font-size": 12, "font-weight": 700, "font-family": "system-ui,sans-serif", dy: 4, dx: 8 }));
    t.textContent = k; t.style.display = "none"; endTexts[k] = t;
  });

  // Legend (right margin)
  [
    { label: "Susceptible", color: "#3b82f6" },
    { label: "Infected",    color: "#dc2626" },
    { label: "Recovered",   color: "#16a34a" },
  ].forEach((d, i) => {
    const ly = tmg.t + 14 + i * 20;
    const lx = TW - tmg.r + 14;
    tsSvg.appendChild(el("line", { x1: lx, x2: lx + 18, y1: ly, y2: ly, stroke: d.color, "stroke-width": 3 }));
    const t = tsSvg.appendChild(el("text", { x: lx + 24, y: ly + 4, fill: "#475569", "font-size": 11, "font-family": "system-ui,sans-serif" }));
    t.textContent = d.label;
  });

  wrapper.appendChild(tsSvg);

  // ══════════════════════════════════════════════════════
  // 7. DOT GEOMETRY  (small dots so we can pack many)
  // ══════════════════════════════════════════════════════
  const DOT_R = 3, DOT_GAP = 1.6;
  const STRIDE = 2 * DOT_R + DOT_GAP;
  const PADX = 10, PAD_TOP = 50, PAD_BOT = 18;
  const innerW = boxW - 2 * PADX;
  const innerH = boxH - PAD_TOP - PAD_BOT;
  const COLS = Math.floor(innerW / STRIDE);
  const ROWS = Math.floor(innerH / STRIDE);
  const MAX_PER_BOX = COLS * ROWS;
  const Y0 = boxY + PAD_TOP + DOT_R;

  // ══════════════════════════════════════════════════════
  // 8. SIMULATION STATE  (deterministic — S, I, R as continuous values)
  // ══════════════════════════════════════════════════════
  let S = 0, I = 0, R = 0;
  let day = 0;
  let history = [];          // {t, S, I, R}
  let autoId = 0, running = false;
  let totalN = 0;

  // Round S, I, R to integers that still sum to N (largest-remainder method),
  // so the dot counts in the boxes are always exactly N together.
  function getCounts() {
    const vals = [S, I, R];
    const floors = vals.map(Math.floor);
    let remaining = totalN - (floors[0] + floors[1] + floors[2]);
    const fracs = vals.map((v, idx) => ({ idx, frac: v - floors[idx] }))
                      .sort((a, b) => b.frac - a.frac);
    const out = floors.slice();
    for (let k = 0; k < remaining && k < 3; k++) out[fracs[k].idx]++;
    return { S: out[0], I: out[1], R: out[2] };
  }

  // ══════════════════════════════════════════════════════
  // 9. RENDER  (no flying — dots only appear / disappear in place)
  // ══════════════════════════════════════════════════════
  function render(animate = true) {
    const cnt = getCounts();
    counts.S.textContent = cnt.S;
    counts.I.textContent = cnt.I;
    counts.R.textContent = cnt.R;

    for (const c of compartments) {
      const n = cnt[c.key];
      const visible = Math.min(n, MAX_PER_BOX);
      const data = [];
      for (let i = 0; i < visible; i++) {
        const row = Math.floor(i / COLS), col = i % COLS;
        data.push({
          i,
          x: c.x + PADX + DOT_R + col * STRIDE,
          y: Y0 + row * STRIDE,
        });
      }
      overflow[c.key].textContent = n > MAX_PER_BOX ? "+" + (n - MAX_PER_BOX) + " hidden" : "";

      // Index-keyed join: dots that were already present keep their circle,
      // dots that no longer fit fade out, new dots fade in at the next free slot.
      const sel = d3.select(dotG[c.key]).selectAll("circle").data(data, d => d.i);

      if (animate) {
        sel.exit()
          .interrupt()
          .transition().duration(320).ease(d3.easeCubicIn)
          .attr("r", 0).style("opacity", 0)
          .remove();

        sel.enter().append("circle")
          .attr("cx", d => d.x).attr("cy", d => d.y)
          .attr("fill", c.color).attr("r", 0).style("opacity", 0)
          .transition().delay(220).duration(380).ease(d3.easeCubicOut)
          .attr("r", DOT_R).style("opacity", 1);
      } else {
        sel.exit().remove();
        sel.enter().append("circle")
          .attr("cx", d => d.x).attr("cy", d => d.y)
          .attr("fill", c.color).attr("r", DOT_R).style("opacity", 1);
      }
    }

    // Stats line
    const beta = SL.beta.val(), gamma = SL.gamma.val();
    const ended = I < 0.5 && day > 0;
    dayLabel.textContent = ended ? `Day ${day} — Epidemic ended` : `Day ${day}`;
    r0Label.textContent  = `R₀ = β / γ = ${(beta / gamma).toFixed(2)}`;
    popLabel.textContent = `N = ${totalN}`;

    drawTimeSeries();
  }

  // ══════════════════════════════════════════════════════
  // 10. TIME-SERIES DRAW
  // ══════════════════════════════════════════════════════
  let tsHiX = 30, tsHiY = 100, tsSx, tsSy;

  function niceCeil(n) {
    if (n <= 100) return Math.ceil(n / 20) * 20;
    if (n <= 500) return Math.ceil(n / 50) * 50;
    return Math.ceil(n / 100) * 100;
  }

  function updateAxes() {
    tsHiX = Math.max(20, Math.ceil((day + 5) / 10) * 10);
    tsHiY = niceCeil(totalN || 100);

    tsSx = t => tmg.l + (t / tsHiX) * tcw;
    tsSy = v => tmg.t + tch - (v / tsHiY) * tch;

    for (let i = 0; i <= 5; i++) {
      const v = (tsHiY / 5) * i, yy = tsSy(v);
      gridLines[i].setAttribute("x1", tmg.l);
      gridLines[i].setAttribute("x2", TW - tmg.r);
      gridLines[i].setAttribute("y1", yy);
      gridLines[i].setAttribute("y2", yy);
      yLabels[i].setAttribute("x", tmg.l - 8);
      yLabels[i].setAttribute("y", yy + 4);
      yLabels[i].textContent = Math.round(v);
    }

    xAx.setAttribute("x1", tmg.l);
    xAx.setAttribute("x2", TW - tmg.r);
    xAx.setAttribute("y1", tsSy(0));
    xAx.setAttribute("y2", tsSy(0));
    yAx.setAttribute("x1", tmg.l);
    yAx.setAttribute("x2", tmg.l);
    yAx.setAttribute("y1", tmg.t);
    yAx.setAttribute("y2", tmg.t + tch);
    xLbl.setAttribute("x", tmg.l + tcw / 2);

    const tsStep = tsHiX > 100 ? 20 : tsHiX > 50 ? 10 : tsHiX > 20 ? 5 : 2;
    let ti = 0;
    for (let x = 0; x <= tsHiX && ti < 24; x += tsStep) {
      const xx = tsSx(x);
      xTickPool[ti].ln.setAttribute("x1", xx);
      xTickPool[ti].ln.setAttribute("x2", xx);
      xTickPool[ti].ln.setAttribute("y1", tsSy(0));
      xTickPool[ti].ln.setAttribute("y2", tsSy(0) + 5);
      xTickPool[ti].ln.style.display = "";
      xTickPool[ti].t.setAttribute("x", xx);
      xTickPool[ti].t.setAttribute("y", tsSy(0) + 5);
      xTickPool[ti].t.textContent = x;
      xTickPool[ti].t.style.display = "";
      ti++;
    }
    for (; ti < 24; ti++) {
      xTickPool[ti].ln.style.display = "none";
      xTickPool[ti].t.style.display = "none";
    }
  }

  function drawTimeSeries() {
    if (history.length === 0) {
      pathS_.setAttribute("d", ""); pathI_.setAttribute("d", ""); pathR_.setAttribute("d", "");
      nowLine.style.display = "none";
      ['S','I','R'].forEach(k => { endDots[k].style.display = "none"; endTexts[k].style.display = "none"; });
      return;
    }

    updateAxes();

    let ds = "", di = "", dr = "";
    for (let i = 0; i < history.length; i++) {
      const h = history[i];
      const x = tsSx(h.t).toFixed(1);
      const cmd = i === 0 ? "M" : "L";
      ds += cmd + x + "," + tsSy(h.S).toFixed(1);
      di += cmd + x + "," + tsSy(h.I).toFixed(1);
      dr += cmd + x + "," + tsSy(h.R).toFixed(1);
    }
    pathS_.setAttribute("d", ds);
    pathI_.setAttribute("d", di);
    pathR_.setAttribute("d", dr);

    if (day > 0) {
      const xNow = tsSx(day);
      nowLine.setAttribute("x1", xNow);
      nowLine.setAttribute("x2", xNow);
      nowLine.setAttribute("y1", tmg.t);
      nowLine.setAttribute("y2", tmg.t + tch);
      nowLine.style.display = "";
    } else {
      nowLine.style.display = "none";
    }

    const last = history[history.length - 1];
    ['S','I','R'].forEach(k => {
      const x = tsSx(last.t), y = tsSy(last[k]);
      endDots[k].setAttribute("cx", x);
      endDots[k].setAttribute("cy", y);
      endDots[k].style.display = "";
      endTexts[k].setAttribute("x", x);
      endTexts[k].setAttribute("y", y);
      endTexts[k].style.display = "";
    });
  }

  // ══════════════════════════════════════════════════════
  // 11. STEP / INIT  (deterministic discrete-time SIR, Δt = 1 day)
  // ══════════════════════════════════════════════════════
  //
  //   ΔS = −β · S · I / N      (people leaving S to I)
  //   ΔI =  β · S · I / N − γ · I
  //   ΔR =                       γ · I
  //
  // S, I, R are tracked as floats so the dynamics stay smooth even when
  // dot counts are small; we only round when drawing the dots.
  function initSim() {
    clearInterval(autoId); running = false; btnAuto.setText("⏩  Auto");
    S = SL.s0.val(); I = SL.i0.val(); R = SL.r0.val();
    totalN = S + I + R;
    day = 0;
    history = [{ t: 0, S, I, R }];
    render(false);
  }

  function stepSim() {
    if (I < 0.5 && day > 0) return;

    const beta = SL.beta.val(), gamma = SL.gamma.val(), N = totalN;
    const newInf = beta * S * I / N;   // S → I flow this day
    const newRec = gamma * I;          // I → R flow this day

    S = Math.max(0, S - newInf);
    I = Math.max(0, I + newInf - newRec);
    R = Math.max(0, R + newRec);

    day++;
    history.push({ t: day, S, I, R });
    render(true);
  }

  // ══════════════════════════════════════════════════════
  // 12. EVENT WIRING
  // ══════════════════════════════════════════════════════
  function onStep() { stepSim(); }
  function onAuto() {
    if (running) {
      clearInterval(autoId); running = false; btnAuto.setText("⏩  Auto"); return;
    }
    running = true; btnAuto.setText("⏸  Pause");
    autoId = setInterval(() => {
      if (I < 0.5 && day > 0) {
        clearInterval(autoId); running = false; btnAuto.setText("⏩  Auto"); return;
      }
      stepSim();
    }, 800);
  }
  function onReset() { initSim(); }

  btnStep.el.addEventListener("click", onStep);
  btnAuto.el.addEventListener("click", onAuto);
  btnReset.el.addEventListener("click", onReset);

  function onParam() {
    SL.s0.sync(); SL.i0.sync(); SL.r0.sync(); SL.beta.sync(); SL.gamma.sync();
    if (!running) {
      initSim();
    } else {
      // β / γ can be tuned mid-run — refresh the R₀ readout immediately.
      const beta = SL.beta.val(), gamma = SL.gamma.val();
      r0Label.textContent = `R₀ = β / γ = ${(beta / gamma).toFixed(2)}`;
    }
  }
  SL.s0.input.addEventListener("input", onParam);
  SL.i0.input.addEventListener("input", onParam);
  SL.r0.input.addEventListener("input", onParam);
  SL.beta.input.addEventListener("input", onParam);
  SL.gamma.input.addEventListener("input", onParam);

  // ══════════════════════════════════════════════════════
  // 13. BOOT + CLEANUP
  // ══════════════════════════════════════════════════════
  initSim();

  invalidation.then(() => {
    clearInterval(autoId);
    btnStep.el.removeEventListener("click", onStep);
    btnAuto.el.removeEventListener("click", onAuto);
    btnReset.el.removeEventListener("click", onReset);
    SL.s0.input.removeEventListener("input", onParam);
    SL.i0.input.removeEventListener("input", onParam);
    SL.r0.input.removeEventListener("input", onParam);
    SL.beta.input.removeEventListener("input", onParam);
    SL.gamma.input.removeEventListener("input", onParam);
    history = [];
  });

  wrapper.value = {};
  return wrapper;
})()
```

## Density or frequency-dependent

Both models compute the **force of infection** $\lambda$ on a susceptible the same way — the **effective contact rate** times the prevalence $I/N$:

$$\lambda \;=\; (\text{effective contact rate}) \;\times\; \frac{I}{N}$$

They differ only in how the effective contact rate is defined:

-   **Density-dependent**: the rate scales with crowd size, $\beta N$. The $N$ cancels the $/N$ from prevalence, so $\lambda = \beta I$. Fits aerosol contacts (e.g. COVID-19).
-   **Frequency-dependent**: the rate is a constant, $\beta$. The $/N$ has nothing to cancel, so $\lambda = \beta I / N$. Fits sexually-transmitted diseases.

```{ojs}
//| echo: false

viewof df_compare = (() => {
  // ══════════════════════════════════════════════════════
  // 1. WRAPPER + STYLE
  // ══════════════════════════════════════════════════════
  const wrapper = document.createElement("div");
  wrapper.style.cssText = `display:flex;flex-direction:column;align-items:center;width:100%;max-width:960px;margin:0 auto;`;
  wrapper.appendChild(injectStyle());

  // ── Geometry & timing ──
  const ROOM_W = 350, ROOM_H = 260;
  const CENTER_X = ROOM_W / 2, CENTER_Y = ROOM_H / 2;
  const DOT_R = 3, FOCAL_R = 7;
  const PREVALENCE = 0.30;     // ~30% of others are infected (red)
  const DD_RADIUS = 65;        // physical reach of the focal (DD)
  const FD_PARTNERS = 5;       // number of fixed closest contacts (FD)
  const PULSE_INTERVAL = 1700;
  const PULSE_DURATION = 750;

  // β for the DD model: probability a uniformly-placed dot lands inside
  // the reach circle = (reach area) / (room area). The displayed contact
  // rate β·N depends only on N (not on where the focal happens to be).
  const BETA_DD = (Math.PI * DD_RADIUS * DD_RADIUS) / (ROOM_W * ROOM_H);

  // ══════════════════════════════════════════════════════
  // 2. SLIDER (N is the only knob)
  // ══════════════════════════════════════════════════════
  const SL = {};
  SL.N = createSlider("N (people sharing the same fixed area)", 30, 500, 10, 200, "#475569", "gray");

  const sliderRow = document.createElement("div");
  sliderRow.style.cssText = "display:flex;justify-content:center;width:100%;margin-bottom:18px;";
  const sliderBox = document.createElement("div");
  sliderBox.style.cssText = "width:100%;max-width:520px;";
  sliderBox.appendChild(SL.N.el);
  sliderRow.appendChild(sliderBox);
  wrapper.appendChild(sliderRow);

  // ══════════════════════════════════════════════════════
  // 3. SVG + MATH HELPERS
  // ══════════════════════════════════════════════════════
  const NS = "http://www.w3.org/2000/svg";
  function el(tag, a) {
    const e = document.createElementNS(NS, tag);
    if (a) for (const [k, v] of Object.entries(a)) e.setAttribute(k, v);
    return e;
  }

  // Render numerator-over-denominator as a real stacked fraction (so I/N
  // looks like math, not text division).
  function frac(num, denom) {
    return `<span style="display:inline-flex;flex-direction:column;vertical-align:middle;text-align:center;line-height:1.05;font-size:0.92em;font-style:italic;margin:0 1px;">` +
           `<span style="border-bottom:1px solid currentColor;padding:0 4px;">${num}</span>` +
           `<span style="padding:0 4px;">${denom}</span>` +
           `</span>`;
  }

  // ══════════════════════════════════════════════════════
  // 4. TWO SIDE-BY-SIDE PANELS  (HTML for the chrome, SVG for the room)
  // ══════════════════════════════════════════════════════
  const panels = [
    {
      key: 'DD', label: 'DENSITY-DEPENDENT', color: '#d97706',
      contactExpr: '<em>β</em> · <em>N</em>',
      // Step-3 algebra with strikethrough on the cancelling N's
      infMathHTML: `<em>β</em> · <s style="color:#dc2626;text-decoration-thickness:2px"><em>N</em></s> · ${frac('I', '<s style="color:#dc2626;text-decoration-thickness:2px">N</s>')} &nbsp;=&nbsp; <strong><em>β</em> · <em>I</em></strong>`,
      cancelNote: () => `The <strong style="color:#dc2626"><em>N</em></strong> in the contact rate <em>cancels</em> the <strong style="color:#dc2626"><em>N</em></strong> in ${frac('I', 'N')} — only <em>I</em> remains in the equation.`,
      eqnHTML: `${frac('dS', 'dt')} = −<em>β</em> · <em>S</em> · <em>I</em>`,
    },
    {
      key: 'FD', label: 'FREQUENCY-DEPENDENT', color: '#0891b2',
      contactExpr: '<em>β</em>',
      infMathHTML: `<em>β</em> · ${frac('I', '<strong style="color:#7c3aed">N</strong>')} &nbsp;=&nbsp; <strong><em>β</em> · ${frac('I', '<span style="color:#7c3aed">N</span>')}</strong>`,
      cancelNote: () => `The contact rate has no <em>N</em> to cancel — the <strong style="color:#7c3aed">${frac('1', 'N')}</strong> from ${frac('I', 'N')} stays in the equation.`,
      eqnHTML: `${frac('dS', 'dt')} = −<em>β</em> · <em>S</em> · ${frac('I', 'N')}`,
    },
  ];

  const panelRow = document.createElement("div");
  panelRow.style.cssText = "display:flex;gap:20px;width:100%;";

  const ref = {};   // per-panel handles for animation

  for (const p of panels) {
    const panel = document.createElement("div");
    panel.style.cssText = `flex:1;display:flex;flex-direction:column;border-radius:12px;overflow:hidden;border:2px solid ${p.color};background:#fff;`;

    // Header — just the panel label.
    const header = document.createElement("div");
    header.style.cssText = `padding:14px;background:${p.color}14;text-align:center;color:${p.color};font-size:13px;font-weight:800;letter-spacing:0.6px;`;
    header.textContent = p.label;
    panel.appendChild(header);

    // SVG room
    const svgEl = el("svg");
    svgEl.setAttribute("viewBox", `0 0 ${ROOM_W} ${ROOM_H}`);
    svgEl.style.cssText = `width:100%;display:block;background:#fafbfc;`;

    // For DD: faint dashed reach circle (moves with the focal blue dot).
    // Drawn in the focal's blue so it visually belongs to the focal — it's
    // *your* reach, not a property of the panel.
    let reachCircle = null;
    if (p.key === 'DD') {
      reachCircle = svgEl.appendChild(el("circle", {
        cx: CENTER_X, cy: CENTER_Y, r: DD_RADIUS,
        fill: '#3b82f6', "fill-opacity": 0.07,
        stroke: '#3b82f6', "stroke-width": 1, "stroke-dasharray": "4 3", "stroke-opacity": 0.45,
      }));
    }

    // For FD: persistent dashed lines to the current closest-k partners
    const partnerLinesG = (p.key === 'FD') ? svgEl.appendChild(el("g")) : null;

    const dotsG    = svgEl.appendChild(el("g"));   // population (red infected + gray susceptible)
    const overlayG = svgEl.appendChild(el("g"));   // pulses + lines

    // Focal "YOU" — a single susceptible (blue, draggable)
    const focal = svgEl.appendChild(el("circle", {
      cx: CENTER_X, cy: CENTER_Y, r: FOCAL_R,
      fill: '#3b82f6', stroke: 'white', "stroke-width": 1.5,
    }));
    const focalLabel = svgEl.appendChild(el("text", {
      x: CENTER_X, y: CENTER_Y - FOCAL_R - 6,
      "text-anchor": "middle", fill: '#3b82f6', "font-size": 9, "font-weight": 800,
      "font-family": "system-ui,sans-serif", "pointer-events": "none",
    }));
    focalLabel.style.letterSpacing = "0.5px";
    focalLabel.textContent = 'YOU';

    panel.appendChild(svgEl);

    // Math display: contact rate × I/N = infection rate, with cancellation
    const math = document.createElement("div");
    math.style.cssText = `padding:14px 16px;border-top:1px solid #e2e8f0;font-size:12px;color:#334155;display:flex;flex-direction:column;gap:10px;background:#fff;`;
    math.innerHTML = `
      <div>
        <div style="color:#64748b;font-size:10.5px;text-transform:uppercase;letter-spacing:0.3px;margin-bottom:3px;">Effective contact rate</div>
        <div style="display:flex;justify-content:space-between;align-items:baseline;font-family:'Latin Modern Math','STIX Two Math',serif;">
          <span style="color:#1e293b;">${p.contactExpr}</span>
          <span style="color:${p.color};font-weight:700;font-family:'SF Mono',monospace;">≈ <span class="rate-val">--</span></span>
        </div>
      </div>
      <div>
        <div style="color:#64748b;font-size:10.5px;text-transform:uppercase;letter-spacing:0.3px;margin-bottom:3px;">× <em>P</em>(contact is <em>I</em>)</div>
        <div style="display:flex;justify-content:space-between;align-items:baseline;">
          <span style="color:#1e293b;">${frac('I', 'N')}</span>
          <span style="color:#7c3aed;font-weight:700;font-family:'SF Mono',monospace;">= ${PREVALENCE.toFixed(2)}</span>
        </div>
      </div>
      <div style="border-top:1px dashed #e2e8f0;padding-top:8px;">
        <div style="color:#64748b;font-size:10.5px;text-transform:uppercase;letter-spacing:0.3px;margin-bottom:6px;">Force of infection</div>
        <div style="font-family:'Latin Modern Math','STIX Two Math',serif;font-size:14px;text-align:center;color:#1e293b;line-height:1.7;">${p.infMathHTML}</div>
        <div style="text-align:center;margin-top:6px;font-family:'SF Mono',monospace;color:${p.color};font-weight:800;font-size:14px;">
          ≈ <span class="inf-val">--</span>
        </div>
      </div>
    `;
    panel.appendChild(math);

    // Mechanism note. `flex-grow:1` lets this section absorb any height
    // difference between the two panels — whichever panel has shorter
    // content stretches its (tinted) note instead of leaving a white gap
    // above the equation footer. The inner text is vertically centred so
    // it sits nicely in the expanded area.
    const mech = document.createElement("div");
    mech.style.cssText = `padding:12px 16px;background:${p.color}10;border-top:1px solid #e2e8f0;flex-grow:1;display:flex;align-items:center;`;
    mech.innerHTML = `<div style="font-size:11px;color:#475569;line-height:1.55;">${p.cancelNote()}</div>`;
    panel.appendChild(mech);

    // Equation footer
    const eqn = document.createElement("div");
    eqn.style.cssText = `padding:11px 14px;background:${p.color}1c;text-align:center;border-top:1px solid ${p.color}40;`;
    eqn.innerHTML = `<div style="font-size:14px;color:#1e293b;font-family:'Latin Modern Math','STIX Two Math',serif;">${p.eqnHTML}</div>`;
    panel.appendChild(eqn);

    panelRow.appendChild(panel);

    ref[p.key] = {
      svgEl, dotsG, overlayG, focal, focalLabel, partnerLinesG, reachCircle,
      rateVal: math.querySelector('.rate-val'),
      infVal:  math.querySelector('.inf-val'),
    };
  }
  wrapper.appendChild(panelRow);

  // ══════════════════════════════════════════════════════
  // 5. CAPTION
  // ══════════════════════════════════════════════════════
  const note = document.createElement("div");
  note.style.cssText = `font-size:12px;color:#64748b;margin-top:14px;text-align:center;line-height:1.7;max-width:760px;`;
  note.innerHTML = `Both rooms hold the <em>same fixed area</em>. The blue dot is "you" — one focal susceptible — and ~30% of the others are red (infected). Every ${(PULSE_INTERVAL/1000).toFixed(1)}s "you" make contact with people: <strong style="color:#d97706">DD</strong> reaches everyone in the dashed circle (denser room ⇒ more contacts), <strong style="color:#0891b2">FD</strong> reaches your ${FD_PARTNERS} closest neighbours (dashed teal lines) regardless of crowd size. Drag the slider, or <strong>grab the blue dot</strong> and walk through the crowd. Then read the math underneath each panel: it's the same recipe in both — (contact rate) × (chance contact is with <em>I</em>) = infection rate. The only difference is whether the contact rate has an <em>N</em> in it that cancels with the <em>N</em> in <em>I/N</em>.`;
  wrapper.appendChild(note);

  // ══════════════════════════════════════════════════════
  // 6. POPULATION (focal at index 0, ~30% of others infected)
  // ══════════════════════════════════════════════════════
  let positions = [];   // index 0 is the focal blue dot
  let fdPartners = [];  // indices into positions for current closest-k partners

  let rngSeed = 12345;
  function rng() {
    rngSeed = (rngSeed * 16807) % 2147483647;
    return (rngSeed - 1) / 2147483646;
  }

  function regeneratePositions(N) {
    rngSeed = 12345;
    positions = [{ x: CENTER_X, y: CENTER_Y, infected: false }];
    const margin = DOT_R + 2;
    const minDistFromFocal = FOCAL_R + DOT_R + 4;
    for (let i = 1; i < N; i++) {
      let x, y, tries = 0;
      do {
        x = margin + rng() * (ROOM_W - 2 * margin);
        y = margin + rng() * (ROOM_H - 2 * margin);
      } while (Math.hypot(x - CENTER_X, y - CENTER_Y) < minDistFromFocal && ++tries < 25);
      positions.push({ x, y, infected: false });
    }
    // Mark ~30% of others as infected (random subset, deterministic via seed)
    const N_others = positions.length - 1;
    const I_count = Math.round(PREVALENCE * N_others);
    const idx = [];
    for (let i = 1; i < positions.length; i++) idx.push(i);
    for (let i = idx.length - 1; i > 0; i--) {
      const j = Math.floor(rng() * (i + 1));
      [idx[i], idx[j]] = [idx[j], idx[i]];
    }
    for (let i = 0; i < I_count; i++) positions[idx[i]].infected = true;
  }

  // FD picks the k *closest* dots to the focal — when the user drags the
  // blue dot toward a cluster, those dots become its partners; drag away
  // and a different set takes over.
  function recomputeFDPartners() {
    const cx = positions[0].x, cy = positions[0].y;
    const dists = [];
    for (let i = 1; i < positions.length; i++) {
      dists.push({ i, d: Math.hypot(positions[i].x - cx, positions[i].y - cy) });
    }
    dists.sort((a, b) => a.d - b.d);
    fdPartners = dists.slice(0, Math.min(FD_PARTNERS, dists.length)).map(x => x.i);
  }

  // ══════════════════════════════════════════════════════
  // 7. RENDER  (dots + persistent FD partner lines + stats)
  // ══════════════════════════════════════════════════════
  function redrawFDPartnerLines() {
    const g = ref.FD.partnerLinesG;
    if (!g) return;
    while (g.firstChild) g.removeChild(g.firstChild);
    const cx = positions[0].x, cy = positions[0].y;
    for (const i of fdPartners) {
      if (i >= positions.length) continue;
      const t = positions[i];
      g.appendChild(el("line", {
        x1: cx, y1: cy, x2: t.x, y2: t.y,
        stroke: '#0891b2', "stroke-width": 1.8, "stroke-dasharray": "4 3", opacity: 0.55,
      }));
    }
  }

  // Sync the focal blue dot (and DD's reach circle) to positions[0] in both panels.
  function syncFocal() {
    const cx = positions[0].x, cy = positions[0].y;
    for (const p of panels) {
      ref[p.key].focal.setAttribute("cx", cx);
      ref[p.key].focal.setAttribute("cy", cy);
      ref[p.key].focalLabel.setAttribute("x", cx);
      ref[p.key].focalLabel.setAttribute("y", cy - FOCAL_R - 6);
      if (ref[p.key].reachCircle) {
        ref[p.key].reachCircle.setAttribute("cx", cx);
        ref[p.key].reachCircle.setAttribute("cy", cy);
      }
    }
  }

  // Drag handler on the focal blue dot. Both panels share positions[0], so
  // dragging either moves both. The focal can't be dragged onto another dot
  // (it gets pushed off), so every closest-k partner stays visually distinct.
  function attachFocalDrag(panelKey) {
    const focal = ref[panelKey].focal;
    focal.style.cursor = "grab";
    d3.select(focal).call(d3.drag()
      .on("start", function(event) {
        event.sourceEvent.stopPropagation();
        this.style.cursor = "grabbing";
        for (const p of panels) {
          d3.select(ref[p.key].focal).interrupt().attr("r", FOCAL_R + 1.5);
        }
      })
      .on("drag", function(event) {
        const [x, y] = d3.pointer(event.sourceEvent, this.ownerSVGElement);
        const margin = FOCAL_R + 2;
        let nx = Math.max(margin, Math.min(ROOM_W - margin, x));
        let ny = Math.max(margin, Math.min(ROOM_H - margin, y));
        // Push focal away from any dot it would overlap. A few iterations
        // handle the case where pushing off one dot lands on another.
        const minD = FOCAL_R + DOT_R + 1;
        for (let it = 0; it < 4; it++) {
          let pushed = false;
          for (let i = 1; i < positions.length; i++) {
            const dx = nx - positions[i].x, dy = ny - positions[i].y;
            const d = Math.hypot(dx, dy);
            if (d < minD) {
              const ang = (d > 0.01) ? Math.atan2(dy, dx) : (rng() * Math.PI * 2);
              nx = positions[i].x + Math.cos(ang) * minD;
              ny = positions[i].y + Math.sin(ang) * minD;
              pushed = true;
            }
          }
          if (!pushed) break;
        }
        nx = Math.max(margin, Math.min(ROOM_W - margin, nx));
        ny = Math.max(margin, Math.min(ROOM_H - margin, ny));
        positions[0].x = nx;
        positions[0].y = ny;
        syncFocal();
        recomputeFDPartners();
        redrawFDPartnerLines();
        updateStats();
      })
      .on("end", function() {
        this.style.cursor = "grab";
        for (const p of panels) {
          d3.select(ref[p.key].focal).attr("r", FOCAL_R);
        }
      })
    );
  }

  function renderDots() {
    syncFocal();   // re-centre focal + reach circle when N changes
    for (const p of panels) {
      const { dotsG, overlayG } = ref[p.key];
      while (dotsG.firstChild)    dotsG.removeChild(dotsG.firstChild);
      while (overlayG.firstChild) overlayG.removeChild(overlayG.firstChild);

      for (let i = 1; i < positions.length; i++) {
        const dot = el("circle", {
          cx: positions[i].x, cy: positions[i].y, r: DOT_R,
          fill: positions[i].infected ? '#64748b' : '#cbd5e1',
        });
        dot.dataset.idx = i;
        dotsG.appendChild(dot);
      }
    }
  }

  function updateStats() {
    // Theoretical contact rates — depend only on N, not on where the
    // focal happens to be standing right now.
    const N = positions.length;
    const ddRate = BETA_DD * N;          // β · N  (DD)
    const fdRate = FD_PARTNERS;          // β       (FD, constant)
    ref.DD.rateVal.textContent = ddRate.toFixed(1);
    ref.DD.infVal.textContent  = (ddRate * PREVALENCE).toFixed(1);
    ref.FD.rateVal.textContent = String(fdRate);
    ref.FD.infVal.textContent  = (fdRate * PREVALENCE).toFixed(1);
  }

  // ══════════════════════════════════════════════════════
  // 8. PULSE ANIMATIONS
  // ══════════════════════════════════════════════════════
  function flashFocal(panelKey) {
    const focal = ref[panelKey].focal;
    d3.select(focal).interrupt()
      .transition().duration(120).attr("r", FOCAL_R + 2.5)
      .transition().duration(280).attr("r", FOCAL_R);
  }

  // Flash a contacted dot — just an enlarge-then-shrink pulse, no colour
  // change. Keeps the room visually quiet; the only motion is the size pop.
  function flashDot(panelKey, i, delay) {
    const dotEl = ref[panelKey].dotsG.querySelector(`[data-idx="${i}"]`);
    if (!dotEl) return;
    d3.select(dotEl).interrupt()
      .transition().delay(delay).duration(120).attr("r", DOT_R + 1.5)
      .transition().duration(450).delay(280).attr("r", DOT_R);
  }

  function pulseDD() {
    flashFocal('DD');
    const { overlayG } = ref.DD;
    const cx = positions[0].x, cy = positions[0].y;

    // Expanding wave from the focal's current position
    const wave = el("circle", {
      cx, cy, r: FOCAL_R,
      fill: 'none', stroke: '#3b82f6', "stroke-width": 2.5, opacity: 0.7,
    });
    overlayG.appendChild(wave);
    d3.select(wave).transition().duration(PULSE_DURATION).ease(d3.easeQuadOut)
      .attr("r", DD_RADIUS).style("opacity", 0).remove();

    // Each in-reach dot flashes when the wave reaches it
    for (let i = 1; i < positions.length; i++) {
      const dist = Math.hypot(positions[i].x - cx, positions[i].y - cy);
      if (dist > DD_RADIUS) continue;
      const delay = (dist / DD_RADIUS) * PULSE_DURATION * 0.7;
      flashDot('DD', i, delay);
    }
  }

  function pulseFD() {
    flashFocal('FD');
    const { overlayG } = ref.FD;
    const cx = positions[0].x, cy = positions[0].y;

    // Lines fan out from the focal to each closest-k partner. Start at the
    // focal's edge (not its centre) so even a partner snug against the focal
    // gets a visible line.
    for (const i of fdPartners) {
      if (i >= positions.length) continue;
      const target = positions[i];
      const dx = target.x - cx, dy = target.y - cy;
      const d  = Math.hypot(dx, dy);
      if (d < 0.5) continue;
      const x1 = cx + (dx / d) * FOCAL_R;
      const y1 = cy + (dy / d) * FOCAL_R;

      const line = el("line", {
        x1, y1, x2: x1, y2: y1,
        stroke: '#3b82f6', "stroke-width": 2.2, opacity: 0.9,
      });
      overlayG.appendChild(line);
      d3.select(line)
        .transition().duration(280).ease(d3.easeQuadOut)
        .attr("x2", target.x).attr("y2", target.y)
        .transition().delay(180).duration(380)
        .style("opacity", 0).remove();

      flashDot('FD', i, 280);
    }
  }

  let pulseTimer = null;
  function tick() { pulseDD(); pulseFD(); }

  // ══════════════════════════════════════════════════════
  // 9. WIRING
  // ══════════════════════════════════════════════════════
  function update() {
    SL.N.sync();
    regeneratePositions(SL.N.val());
    recomputeFDPartners();
    renderDots();
    redrawFDPartnerLines();
    updateStats();
  }

  SL.N.input.addEventListener("input", update);

  update();
  attachFocalDrag('DD');
  attachFocalDrag('FD');
  setTimeout(tick, 250);                          // first pulse soon after mount
  pulseTimer = setInterval(tick, PULSE_INTERVAL); // then every PULSE_INTERVAL ms

  invalidation.then(() => {
    clearInterval(pulseTimer);
    SL.N.input.removeEventListener("input", update);
  });

  wrapper.value = {};
  return wrapper;
})()
```

- Transmission increases as density increases.
- Aerosol contacts (COVID-19).

- Transmission does not increase as density increases.
- Sexually-transmitted diseases.


The two models differ in how a single susceptible "feels" the infectives. Define the **force of infection** $\lambda$ — the rate at which any one susceptible becomes infected:

$$\lambda_{\text{DD}} = \beta I, \qquad \lambda_{\text{FD}} = \beta \frac{I}{N}$$

Hold the prevalence $I/N$ fixed and grow $N$: the DD force of infection grows with $N$ (more bodies in the same space → more contacts), while the FD force is unchanged (each person makes the same number of contacts regardless of crowd size). The same ratio carries through to the basic reproduction number:

$$R_0^{\text{DD}} = \frac{\beta N}{\gamma}, \qquad R_0^{\text{FD}} = \frac{\beta}{\gamma}$$

The two parameterisations agree at exactly one population size; move $N$ away from that calibration point and the predicted epidemics diverge sharply. The illustration below makes this concrete.

## Code

```{r, message=FALSE, warning=FALSE}
library(odin)
library(tidyr)
library(ggplot2)
library(RColorBrewer)
library(patchwork)
```

Below is how we numerically solve the differential equations @eq-sir-ds, @eq-sir-di, @eq-sir-dr in R with the `odin` package.

::: {#lst-sir-det}
```{r, results=FALSE, message=FALSE}
odin_sir <- odin::odin({
  # Derivatives
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(CInc) <- beta * S * I
  
  # Initial conditions
  initial(S) <- S_init
  initial(I) <- I_init
  initial(R) <- R_init
  initial(CInc) <- CInc_init
  
  # Parameters and initial values
  beta <- user(0.004)
  gamma <- user(0.5)
  S_init <- user(999)
  I_init <- user(1)
  R_init <- user(0)
  CInc_init <- user(0)
})

sir_mod <- function(beta, gamma, S0, I0, R0, times) {
  # Set values for a new run
  odin_run <- odin_sir$new(beta = beta, gamma = gamma, S_init = S0, I_init = I0, R_init = R0)
  
  # Run the model
  out <- data.frame(odin_run$run(times))

  # Compute incidence from cumulative incidence
  out$Inc <- c(I0, diff(out$CInc))
  out$CInc <- NULL
  out
}
```

Deterministic SIR model
:::

## Incidence or prevalence

::: callout-important
Be careful of what kind of data we are using to fit the model:

-   **Incidence**: use the $\beta SI$ part.

$$\text{Incidence} = \frac{\text{Number of new cases during a period}}{\text{Population at risk during the same period}}$$

-   **Prevalence**: use the **I compartment**.

$$\text{Prevalence} = \frac{\text{Number of existing cases at time t}}{\text{Total population at time t}}$$
:::

There are two ways to compute incidence.

1.  While writing the differential equations, we tell `odin` to compute the cumulative incidence, then compute the incidence by using `diff()` on the cumulative incidence.

```{r, eval=F}
# Cumulative incidence
deriv(CInc) <- beta * S * I
# Compute incidence from cumulative incidence
out$Inc <- c(I0, diff(out$CInc))
```

2.  In this "closed population", the incidence is the number of individuals moving out of $S$ in each time step. So we can compute it from the $S$ column. Let compare it with the previous method.

```{r}
pred <- sir_mod(beta = 0.004, gamma = 0.5, S0 = 999, I0 = 1, R0 = 0, times = 0:13)

tmp <- pred
tmp$Inc2 <- c(tmp$I[1], -diff(tmp$S))
tmp[,"Inc-Inc2"] <- round(tmp$Inc - tmp$Inc2)
tmp[,c("Inc", "Inc2", "Inc-Inc2")]
```

Now let have a look at the output.

```{r}
head(pred)
```

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "100%"
df_plot <- pivot_longer(pred, cols = S:Inc, names_to = "comp", values_to = "n")

ggplot(df_plot, aes(x = t, y = n, color = comp)) +
  geom_line(linewidth = 1.2) +
  scale_color_brewer(palette = "PuOr", breaks = c("S", "Inc", "I", "R")) +
  labs(color = NULL, y = NULL, x = "Time") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The time to peak of incidence is earlier than prevalence:

-   Incidence reflects the immediate rate of new infections, the peak of incidence is reached when the transmission rate begins to decline.
-   Prevalence accumulates both new and existing cases. Even after the incidence starts to decline, the active cases who were infected before the incidence peak are still there waiting for recovering. Prevalence only starts to decline when the recovery rate surpasses the rate of new infections significantly.

::: callout-important

Normally, public health officials are interested in the **incidence** in the general population (to track disease spread) and the **prevalence** in intensive care units (to ensure it does not exceed ICU bed capacity).

:::

## Sensitivity analysis

The effective contact rate $\beta$ and the recovery rate $\gamma$ both affect whether or not there is an epidemic. The total number of people in the population and the initial number of infected people have no effect on whether there is an epidemic.

Below is a function that take a range of values of parameters and compare the model outputs.

```{r}
#| code-fold: true
sir_sa <- function(vals, parm, outcome, times, init_beta = 0.004, init_gamma = 0.5) {
  pick_comp <- ifelse(outcome == "prev", "I", "Inc")
  plot_lab <- ifelse(outcome == "prev", "Prevalence", "Incidence")
  symb <- ifelse(parm == "beta", "\u03b2", "\u03b3")
  l <- list()
  for(i in 1:length(vals)) {
    if (parm == "beta") {
      pred <- sir_mod(beta = vals[i], gamma = init_gamma, S0 = 999, I0 = 1, R0 = 0, times = times)[, c("t", pick_comp)]
    } else {
      pred <- sir_mod(beta = init_beta, gamma = vals[i], S0 = 999, I0 = 1, R0 = 0, times = times)[, c("t", pick_comp)]
    }
    pred$comp <- paste0(pick_comp, i)
    colnames(pred) <- c("t", "n", "comp")
    l[[i]] <- pred
  }
  # Bind this list into a data frame
  df_plot <- Reduce(rbind, l)
  # Set level I1 < I2 < I3... according to the increase of beta values
  df_plot$comp <- factor(df_plot$comp, levels = c(paste0(pick_comp, 1:length(vals))))
  
  # Plot results
  my_palette <- brewer.pal(9, "Blues")[(9-length(vals)+1):9]
  
  ggplot(df_plot, aes(x = t, y = n, color = comp)) + 
    geom_line(linewidth = 1.2) +
    scale_color_manual(values = my_palette, labels = c(paste0(symb, " = ", vals))) +
    labs(color = NULL, y = plot_lab, x = "Time") +
    theme_minimal()
}
```

### Effective contact rate $\beta$

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 4
#| out-width: "100%"

vals <- seq(0.001, 0.006, 0.001)

p1 <- sir_sa(vals = vals, parm = "beta", outcome = "prev", times = 0:13) + labs(x = NULL)
p2 <- sir_sa(vals = vals, parm = "beta", outcome = "inc", times = 0:13)

p1 / p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

When $\beta$ **decreases**, less susceptible individuals become infected per unit time:

-   Time until peak ***increases***: because the speed of susceptible individuals become infected is slower.
-   Epidemic duration ***increases***: the rate of new infections slows down so it can lead to a longer epidemic period, but depends on the balance between the infection rate and the recovery rate.
-   Outbreak size **decreases**: $R_0 = \frac{\beta}{\gamma}$, so if $\beta$ decreases, $R_0$ **decreases**, each infected individual infect less people on average.
-   **Lower** epidemic peak: because fewer infections occurring simultaneously.

Interventions that decrease $\beta$ cause a lower outbreak size and a lower epidemic peak. This effect is often referred to as **"flattening the curve"**.

### Recovery rate $\gamma$

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 4
#| out-width: "100%"

vals <- seq(0.3, 0.8, 0.1)

p1 <- sir_sa(vals = vals, parm = "gamma", outcome = "prev", times = 0:13) + labs(x = NULL)
p2 <- sir_sa(vals = vals, parm = "gamma", outcome = "inc", times = 0:13)

p1 / p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

When $\gamma$ **decreases**, patients recover more slowly meaning that they remain infectious for a longer period:

-   Outbreak size ***increases***: since $R_0 = \frac{\beta}{\gamma}$, a decrease in $\gamma$ leads to an increase in $R_0$, meaning each infected individual infects more people on average.
-   ***Higher*** epidemic peak: with a higher $R_0$ more individuals are infected after each generation.
-   Time until peak ***may increases***: the infected population takes longer to clear out due to slower recovery, but the exact timing of the peak also depends on the balance between infection and recovery rates.
-   Epidemic duration ***slightly increases***: for same reason as the peak time, the overall duration of the epidemic is prolonged due to slower recovery rates.

## Corollaries

### The threshold phenomenon

From @eq-sir-di:

$$\frac{dI}{dt} = \beta SI - \gamma I = I(\beta S - \gamma)$$

The infection "dies out" when:

$$\frac{dI}{dt} < 0 \Leftrightarrow \beta S - \gamma < 0 \Leftrightarrow S < \frac{\gamma}{\beta}$$

> The proportion of susceptibles in the population must exceed this critical threshold $\frac{\gamma}{\beta}$ for an infection to invade [@keeling2011].

The inverse of the relative removal rate is called the basic reproductive ratio $R_0 = \frac{\beta}{\gamma}$ [@keeling2011]. See @def-r0 for more details.

### Epidemic burnout

Divide @eq-sir-ds by @eq-sir-dr:

$$\frac{dS}{dR} = \frac{-\beta S I}{\gamma I} = -\frac{\beta}{\gamma} S$$

Upon integrating with respect to R, we obtain:

$$S(t) = S(0)e^{-\frac{\beta}{\gamma} R(t)}$$

Given that $e^x > 0 \quad \forall x$, therefore $S(t) > 0$ there will always be some susceptibles in the population who escape infection, leading to an important counter-intuitive conclusion from this model:

> The chain of transmission eventually breaks due to the decline in infectives, not due to a complete lack of susceptibles [@keeling2011].
