# Model vaccination

::: {.callout-note appearance="simple"}
# Acknowledgement

This chapter is based on the teachings of my PhD supervisor, Dr. Marc Choisy, and a lecture given by Asst Prof. Hannah Clapham at MIDSEA Summer School 2024, for which I am deeply grateful.
:::

::: {.callout-note collapse="true"}

# Slides

```{=html}
<div>
<iframe class="slide-deck" style="height: 50vh; width: 100%;" src="slides/vax-mod.html"></iframe>
</div>
```

[{{< fa expand size=fw >}} Full screen](slides/vax-mod.html){.btn .btn-outline-secondary role="button"}

:::

To model vaccination, we need to decide 2 things:

1. What is the vaccine effect? Does it protect against infection, severe disease, transmission, or a combination?

2. How do we model imperfect protection?

## Vaccine effects

### Prevent infection

Vaccine prevents the pathogen from establishing an infection in the body. A vaccinated person is less likely to get infected.

![](img/vax/prv-inf.svg)

Assuming a perfect vaccine against infection, the model can be designed as follows:

![](img/vax/prv-inf-mod.svg)

Vaccinated people $V$ remain uninfected in their lifetime and do not participate in disease transmission.

You can simplify this model by moving susceptible people to the $R$ compartment because they also do not participate in disease transmission.

![](img/vax/prv-inf-mod-2.svg)

### Prevent severe disease

Vaccine reduces the severity of the disease if the vaccinated person does get infected. A vaccinated individual still get infected but experiences a much milder form of the disease.

![](img/vax/prv-dis.svg)

Assuming a perfect vaccine against severe disease, the model can be designed as follows:

![](img/vax/prv-dis-mod.svg)

The infected population is split into a proportion of $p_s$ for severe cases and $1 - p_s$ for mild cases. Vaccinated people $S_v$ still get infected, but will only experience mild disease.

### Prevent transmission

Vaccine reduces the ability of a vaccinated individual, who becomes infected, to transmit the pathogen to others. A vaccinated person still get infected, still get sick, but they are less likely to spread the disease to others.

![](img/vax/prv-trans.svg)

Assuming a perfect vaccine against transmission, the model can be designed as follows:

![](img/vax/prv-trans-mod.svg)

$I$ being the only source of infection that contributes to the force of infection. Vaccinated people $S_v$ still get infected and become $I_v$, but they do not transmit the disease.

## Imperfect protections

What is meant by a vaccine with effectiveness of 80% protection against a specific clinical endpoint? The mechanism of vaccine action is typically modelled in two ways:

### All-or-nothing model

::: {.callout-tip appearance="simple"}
**All-or-nothing** (AoN, also called the **polarized** [@park2023], or **take**) model assumes that **among vaccinated** individuals, a **proportion** $VE_P$ are **completely protected**, while **the remaining fraction** $1 - VE_P$ remains **completely unprotected** [@zachreson2023; @park2023].
:::

An effectiveness of 80% here implies that among vaccinated people, 80% are completely protected, and 20% receive no protection [@worldhealthorganization2013].

![](img/vax/aon.svg)

Assumming a standard SIR model, vaccine protects against **infection** with vaccine effectiveness $VE_P$, $\rho$ represents the vaccination rate:

-   Among the susceptible $S$, $\rho S$ are vaccinated.
-   Among those $\rho S$ who are vaccinated, $\rho VE_P S$ are completely protected and go directly to $R_v$, the remaining is $\rho (1 - VE_P) S$.
-   The remaining $\rho (1 - VE_P) S$ are completely unprotected, and will be infected with force of infection $\lambda(t)$ just like those who are unvaccinated.

![All-or-nothing vaccination model [@park2023]](img/vax/aon-mod.svg){#fig-aon-mod}

$$\begin{align}
\frac{dS}{dt} &= -\lambda(t) S - \rho S \\
\frac{dI}{dt} &= \lambda(t) S - \gamma I \\
\frac{dR}{dt} &= \gamma I \\
\frac{dS_v}{dt} &= \rho (1 - VE_P) S - \lambda(t) S_v \\
\frac{dI_v}{dt} &= \lambda(t) S_v - \gamma I_v \\
\frac{dR_v}{dt} &= \rho VE_P S + \gamma I_v
\end{align}$$

### Leaky model

::: {.callout-tip appearance="simple"}
**Leaky** (or **degree**) model assumes that **all vaccinated** individuals **are partially protected** [@zachreson2023].
:::

An effectiveness of 80% here implies that all vaccinated people have the endpoint of interest reduced by 80% compared to non-vaccinees.

![](img/vax/leaky.svg)

The assumption that no vaccinated people is totally or permanently protected implies one or both of the following [@worldhealthorganization2013]:

-   No amount (titre) of the immune marker is totally protective or, if it is, no individual can maintain that titre for a long period (because of waning or transient immunosuppression)
-   The degree of protection is a function of the level of the immune marker - the simplest explanation being that protection is a function of both the level of the immune marker and the challenge dose.

Assumming a standard SIR model, vaccine protects against **infection** with vaccine effectiveness $VE_L$, $\rho$ represents the vaccination rate.

-   Among susceptible $S$, $\rho S$ are vaccinated.
-   Among those $\rho S$ who are vaccinated, force of infection $\lambda(t)$ is reduced by a factor of $1 - VE_L$.

![Leaky vaccination model [@park2023]](img/vax/leaky-mod.svg){#fig-leaky-mod}

$$\begin{align}
\frac{dS}{dt} &= -\lambda(t) S - \rho S \\
\frac{dI}{dt} &= \lambda(t) S - \gamma I \\
\frac{dR}{dt} &= \gamma I \\
\frac{dS_v}{dt} &= \rho S - (1 - VE_L) \lambda(t) S_v \\
\frac{dI_v}{dt} &= (1 - VE_L) \lambda(t) S_v - \gamma I_v \\
\frac{dR_v}{dt} &= \gamma I_v
\end{align}$$

## Mixed effects

Imperfect vaccines often have a combination of effects: they can protect against infection, severe disease, and transmission at the same time.

![](img/vax/prv-mix.svg)

Assuming a vaccine with leaky effectiveness $VE_L$ against infection and severe disease, the model can be designed as follows:

![](img/vax/prv-inf-dis-mod.svg)

-   Among susceptible $S$, $\rho S$ are vaccinated.
-   Among those $\rho S$ who are vaccinated, force of infection $\lambda(t)$ is reduced by a factor of $1 - VE_L$.
-   Vaccine reduces the proportion of severe disease from $p_s$ to $p_s'$.

## Multiple doses

Some vaccines require multiple doses, like measles (2 doses) or rotavirus (2 doses), with varying effectiveness per dose. Below is the DynaMICE model [@verguet2015] for the 2-dose measles vaccine, assuming AoN effectiveness as in @fig-aon-mod for both doses.

![](img/vax/dynamice.svg)

## Code

### A perfect vaccine against infection

```{r, warning=FALSE, message=FALSE}
library(odin)
library(tidyr)
library(ggplot2)
```

![](img/vax/prv-inf-mod.svg)

```{r, results=FALSE, message=FALSE}
pfvac_ode <- odin({
  # Derivatives
  deriv(S) <- -beta * S * I - rho * S
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(V) <- rho * S
  
  # Initial conditions
  initial(S) <- N_init - I_init
  initial(I) <- I_init
  initial(R) <- R_init
  initial(V) <- V_init
  
  # Parameters and initial values
  beta <- user(8.77e-8)
  gamma <- user(0.2)
  rho <- user(0.005)
  N_init <- user(5700000)
  I_init <- user(1)
  R_init <- user(0)
  V_init <- user(0)
})
```

```{r}
# Initialize model
pfvac_mod <- pfvac_ode$new()
# How long to run
times <- seq(0,300)
# Run the model
pred <- data.frame(pfvac_mod$run(times))
```

```{r}
#| code-fold: true
#| fig-width: 4
#| fig-height: 3
#| out-width: "100%"
df_plot <- pivot_longer(pred, cols = S:V, names_to = "comp", values_to = "n")

ggplot(df_plot, aes(x = t, y = n, color = comp)) +
  geom_line(linewidth = 1.2) +
  scale_color_brewer(palette = "PuOr", breaks = c("S", "I", "R", "V")) +
  labs(color = NULL, y = NULL, x = "Time") +
  theme_minimal() +
  theme(legend.position = "bottom")
```


### AoN model against infection



### Leaky model against infection

